### 2022-04-22 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# article1

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2022-05-18 analysis/paper/paper.Rmd
- first knit test - Introduction progress

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

## Previous research

## Geological setting and chert outcrops

### Geological setting

The Algarve region is located in south Portugal, framed north and east by the Alentejo region and Spain respectively. To the west and south, it is bordered by the Atlantic ocean. Geologically, it is composed of two main geological units: the South Portuguese Zone (SPZ) and the Algarve basin. The SPZ is composed of four "domains" of different lithologies and ages, extending up to Alentejo, although all within the Paleozoic period. It is overlain unconformably by the Mesozoic sedimentary rocks of the basin [\@fernandes_new_2012]. The Algarve basin is one of broader Jurassic outcrops in Portugal [@malchus_microfacies_1988]. It corresponds to the mesozoic-cenozoic sediments which outcrop south of Portugal, from Cape of St. Vicent to Vila Real de St. António. The Algarve basin is associated with the opening of the central Atlantic and with the eventual oceanic crust formation in the western part of the Tethis sea, between the Algarve and North Africa [@terrinha_bacia_2013]. After the sedimentation period between the Triassic and the Sinemurian (Lower Jurassic), the mesozoic basin was divided in two sub-basins (western sub-basin and oriental sub-basin), until the Calovian (Middle Jurassic), when an uplift episode allowed a uniform sedimentation of the basin [@rocha_estudo_1976; @terrinha_bacia_2013]. The existence of two sub-basins and a structural higher area between them, allied with the expansion and retraction of the seas allowed a variation of sedimentation environments, ranging between external platform and internal platform, and continental, hemipelagic, or deep marine environments [@terrinha_bacia_2013]. It was this variance in deposition environments that created the variety of Paleozoic sedimentary facies, with moments of more or less homogeneity throughout this period.

The Algarve basin is of essential importance for the existence of chert in the Algarve region, since it is in the Paleozoic sediments where chert nodules can be found. Paleozoic sedimentation in the basin starts in the Triassic, although the identification of chert nodules or beds is limited to the Lower to Upper Jurassic. It is at the start of the Lower Jurassic, during the Lower Pliesbachian (also named Carixian) that the Algarve basin started to separate into two sub-basins, until the Calovian (Middle Jurassic), when an uplift episode allowed a uniform sedimentation of the basin [@rocha_estudo_1976; @terrinha_bacia_2013]. During the Lower Pliesbachian, the sediments in the western sub-basin can be described as marine, of external platform, while the sediments of the eastern sub-basin are of an internal platform. The transition between the Lower Jurassic to the Middle Jurassic shows a hiatus, as there are no known sediments from the Toarcian to the Aaelinian in the basin, the sedimentation restarting only after the latter age. During the Upper Jurassic the lithofacies variation became more prominent, especially during the Upper Oxfordian-Lower Kimmeridgian, with a certain uniformity of the lithofacies in all the sub-basins during the Upper Kimmeridgian [@terrinha_bacia_2013].

### Chert outcrops

In general, the presence of chert in the Algarve region and be attributed to limestone or dolomitic limestone layers. They appear with relative frequency, and often as nodules of small to medium dimensions, although thin beds of chert can also be identified in some areas.

As mentioned above, outcrops with chert nodules in the Algarve region can be found in the Paleozoic sediments from the Algarve Basin, more specifically in the Jurassic sediments. Chert nodules can be found within Lower, Middle and Upper Jurassic layers, although restricted to a limited number of layers within the stratigraphy and not present in every geological cut.

Lower Jurassic geological layers with chert nodules can be mostly found in the Carixian-age sediments (also known as Pliensbachian) from the Sagres region, often outcropping in the cliffs by the beaches. These include the chert nodules and beds from Cabo de S. Vicente to Praia do Belixe, but also the small outcrop inland, in Ferrel. The chert nodules occur within limestone, dolomitic limestone and marly limestone. Further east one layer with chert nodules was also identified, but attributed to the Sinemurian-Toarcian, in a formation identified as "Limestones with *Paleodasucladus* and limestones with chert nodules". Described only in the Geologic Map at a 1 200 000 scale however, these chert nodules were characterized as micronodules [\@oliveira_carta_1992].

Middle Jurassic geological layers with chert nodules are only found east of Sagres, in the Albufeira, Faro and Tavira region. These layers were attributed to the Malhão Formation, from the Aalenian-Bajocian age. In general, this formation was described as carbonated, marine and thick, and chert can be found in two different layers: monogenic conglomerates with micritic limestone intercalations with chert beds and nodules; microcrystalline limestones with chert nodules (pel-biomicrite). Another chert outcrop was identified in the Geologic Map at a 1 200 000 scale, attributed to the Bajocian-Batonian age and named "Alagoa conglomerates", with partially dolomitic limestones and chert nodules.

Finally, Upper Jurassic sediments have been identified also in the Albufeira, Faro and Tavira area. The Jordana Formation was attributed to the Kimmeridgian/Kimmeridgian-Portlandian, and is characterized by dark-gray limestones, with frequent secondary silifications with abundant fossil fragments. In the Tavira area these layers were instead attributed to the Oxfordian-Kimmeridgian age.

# Materials and methods

## Fieldwork and geological samples

To obtain geological samples several known chert outcrops were visited during August of 2021 and April of 2022, to prospect the Southwest and Southeast portion of the Algarve region (respectively). The first step of fieldwork included the review of previously known research, including geological maps, archaeological prospection works and raw material studies. These were followed as a reference guide for recovering samples. In the field, several chert nodules were collected within one single outcrop, in order to obtain as much variability as possible. The surrounding areas to the outcrops were also surveyed to understand other possible outcrops or secondary deposition environments. All samples were registered with resource to an android app which allows accurate coordinates to be associated with photos and csv data regarding the outcrop characteristics and conditions (i.e. position, abundance, visibility, access, geomorphology, chert morphology and conditions). Individual ID tags were associated with each sample and placed in the corresponding bag.

\*\* outcrops were identified, \*\* in primary and \*\* in secondary deposition. From these, \*\* samples were recovered. The geological samples were taken to the laboratory where they were washed, relabeled to be inserted in the existing collection and analyzed macroscopically.

# Results

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2022-09-20 analysis/paper/paper.Rmd
- First draft of article text without corrections

@@
 author:
-  - Myfirstname Mylastname:
+  - name: Joana Belmiro
     email: fl@oneplace.edu
@@
     correspondence: true
-  - Yourfirstname Yourlastname:
+  - name: João Cascalheira
     email: fl@another.edu
@@
 
+```{r packages, include=FALSE}
+
+library(ggplot2)
+theme_set(theme_bw())
+library(sf)
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(readr)
+library(ggspatial)
+library(ggrepel)
+library(dplyr)
+library(kableExtra)
+library(flextable)
+library(data.table)
+library(tidyr)
+
+```
+
+```{r databases, include=FALSE}
+
+petrography <- read_csv(here::here('analysis/data/raw_data/PetrographyDB.csv')) # Reading the petrography data
+chertfield <- read_csv(here::here('analysis/data/raw_data/chert_field.csv')) # Reading the field data
+
+fielddata <- chertfield %>% # Cleaning the Field data
+  mutate(Geography = case_when(long <= -8.2 ~ "Ocidental",
+                               long > -8.2 ~ "Oriental")) %>% 
+  filter(`Raw Material` == "Chert") %>% 
+  rename(SampleID = NewCode)
+
+```
+
+```{r map setup, include=FALSE}
+
+sf::sf_use_s2(FALSE)
+world <- ne_countries(scale = "medium", returnclass = "sf")
+#europe <- world[world$continent == "Europe",]
+europe_cropped <- st_crop(world, xmin = -15, xmax = -7.5,
+                                    ymin = 35, ymax = 37.6)
+
+```
+
+```{r map, echo=FALSE, include=FALSE}
+
+ggplot(data = europe_cropped) +
+  geom_sf() +
+  geom_point(data = chertfield, aes(x = long, y = lat, shape = State, color = State), size = 4) +
+  #geom_label_repel(data = chertsamples, aes(long, lat, label=NewCode), box.padding = 0.5, max.overlaps = 30) +
+  annotation_scale(location = "br", width_hint = 0.5) +
+  annotation_north_arrow(location = "br", which_north = "true", 
+        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
+        style = north_arrow_fancy_orienteering)
+  #coord_sf(xlim = c(long_min, long_max), ylim = c(lat_min,lat_max), expand = FALSE)
+
+```
+
+```{r databasesetup, include=FALSE}
+
+CStable <- chertfield %>% 
+  select(NewCode, Site, `Sample variability`, State, Deposit, Type, Frequence, `State of site`, `Thin-sections`)
+
+```
+
+```{r cstable, echo=FALSE, include=FALSE}
+
+regulartable(CStable) %>% 
+  theme_zebra()
+
+```
+
 # Introduction
 
-Here is a citation [@Marwick2017]
+The South Iberian territory is a key area to understand cultural transitions since the Middle Paleolithic to early Prehistory.
+As a territory located at the tip of the European continent and with a generally warm climate even during the coldest periods which characterize the Pleistocene, South Iberia, and specifically South Portugal, witnessed several moments of crisis, cultural changes and behavioral adaptations.
+An example of this is the occupation of the territory by Neanderthal populations and their consequent disappearance.
+The southern part of Portugal is characterized by various Neanderthal occupations -- NEED STUDIES.
+The several cultural shifts of the Upper Paleolithic (UP) are also represented in the region, in sites which extend from the first Homo Sapiens occupations with the Gravettian up to the Magdalenian, such as the archaeological site of Vale Boi.
+This site, containing one of the most complete UP sequences of the Iberian Peninsula, offers the opportunity to understand the movements of the first groups of Humans to reach the territory, as well as the impact of abrupt climatic change in culture, with the Heinrich Event 2 and the Last Glacial Maximum.
+Finally, the existence of several Mesolithic, Neolithic and Chalcolithic settlements and occupations, the Algarve regions offers the possibility to study the replacement of the last hunter-gatherers of Europe by the first farms, and the subsequent transformations these communities went through.
 
-## Previous research
+\>\> From \*\*40 000\*\* cal BP to 3 000 cal BP, South Portugal has seen moments of crisis and cultural change of different kinds, which gives this territory a large potential for Pre-history archaeology studies and the study of human behavior in a wide chronological range.
+With such a wide chronology, differences in settlement patterns, subsistence techniques, climate, technology and culture, it is important, however, to find a unifying subject that can be studied, analyzed and compared, bringing together several studies and sites to create a more unified story of the evolution of human behavior.
+That subject may be the raw material which was used by Neanderthals, Human hunter-gatherers and farmers - that raw material is chert.
+
+Chert, whenever geologically present (and at times, even when it is not geologically present), has served as a raw material for the production of blanks to more specialized and complex tools.
+As a good quality rock, it was procured from close to far-away locations, transported, knapped, retouched, re-used, heated and sometimes traded, travelling long distances in the hands of different groups which produced them.
+Countless archaeologists and studies have shown the potential of this raw material to understand cultural aspects such as foraging strategies and territory use, mobility, shifts in technological necessities, cultural preferences, changes in raw material visibility and accessibility, or even the establishment, expansion and shrinkage of social networks.
+
+As a raw material, chert has been studied in South Portugal.
+The study of chert bearing outcrops has been a topic of interest for both geologists and archaeologists in the Algarve region, albeit for different reasons.
+Geological literature has been often focused on the processes that formed the seas and landmasses of the Algarve basins, as well as the paleobiology and paleobiogeography of the region.
+Archaeological literature has focused chert sources as a way to predict the existence of settlements, characterize hunter-gatherer culture throughout prehistory, and understand group mobility and social exchange through time.
+
+Geological studies in the Algarve region go back to the 19th century and throughout all of the 20th century, through the creation of geological cartography.
+These first works provided the basis for the geological knowledge of the Algarve basin, and it allowed further studies within the geosciences, as a large number of Doctoral theses.
+
+Archaeological studies focusing on raw materials, and especially chert, have been more scarce.
+In general, lithic technology studies in the Algarve have approached raw materials through a macroscopic methodology, often related to the location of the settlement (i.e., [\@bicho_o\_2003]) or the description of the collection (i.e., [\@cascalheira_tecnologia_2010], [\@marreiros_lithic_2015]).
+The work of Veríssimo (2005) focusing on the occurrence of chert in western Algarve provided the initial basis for comparative studies, maintaining however solely a macroscopic component.
+Finally, the creation of LusoLit (Telmo et al.
+2016) and the collection of samples from the Algarve provided a new leap in the study of chert in the region.
+
+Given the existence of several chert-bearing outcrops in South Portugal, preliminary raw material studies, a region of archaeological importance and the existence of a rich archaeological record, the Algarve stands as a region full of potential for studies of chert and the development of a lithotheque which may serve for regional studies, but also for comparative studies outside, where chert may have been transported and traded, offering the chance to complete the record of exotic cherts which may be coming from the Algarve.
+
+Given this regional potential and opportunity to complete the knowledge of chert in South Portugal, the aim of this paper is to locate, map and characterize the chert sources in South Portugal, of potential social and economic importance to prehistoric communities.
+As a result, it is our objective continue the lithotheque LusoLit hosted at ICArEHB (University of Algarve) and provide all data in an online database.
 
@@
 It is at the start of the Lower Jurassic, during the Lower Pliesbachian (also named Carixian) that the Algarve basin started to separate into two sub-basins, until the Calovian (Middle Jurassic), when an uplift episode allowed a uniform sedimentation of the basin [@rocha_estudo_1976; @terrinha_bacia_2013].
-During the Lower Pliesbachian, the sediments in the western sub-basin can be described as marine, of external platform, while the sediments of the eastern sub-basin are of an internal platform.
+During the Lower Pliensbachian, the sediments in the western sub-basin can be described as marine, of external platform, while the sediments of the eastern sub-basin are of an internal platform.
 The transition between the Lower Jurassic to the Middle Jurassic shows a hiatus, as there are no known sediments from the Toarcian to the Aaelinian in the basin, the sedimentation restarting only after the latter age.
@@
 
-## Fieldwork and geological samples
+To identify the chert outcrops present in South Portugal and understand the chert's characteristics, a macroscopic and petrographic approach was applied to the study of geological samples.
 
-To obtain geological samples several known chert outcrops were visited during August of 2021 and April of 2022, to prospect the Southwest and Southeast portion of the Algarve region (respectively).
+Geological samples were obtained through fieldwork, during August of 2021 (west Algarve) and April to June of 2022 (east Algarve), and included visits to previously known outcrops and prospection of areas identified in geological maps.
 The first step of fieldwork included the review of previously known research, including geological maps, archaeological prospection works and raw material studies.
 These were followed as a reference guide for recovering samples.
-In the field, several chert nodules were collected within one single outcrop, in order to obtain as much variability as possible.
-The surrounding areas to the outcrops were also surveyed to understand other possible outcrops or secondary deposition environments.
+Whenever necessary, samples from previous prospection works were consulted, from the LusoLit lithoteque from ICArEHB at the University of Algarve and the thin-sections at the University of Évora.
+Whenever coordinates or specific locations for known outcrops were available, those were visited and then the surrounding area was prospected to understand the extension of the outcrops and possible secondary deposition outcrops nearby.
+In all other situations, several locations with potential to find outcrops were selected from polygons where formations with chert were known to outcrop.
+
+Samples were collected whenever possible, focusing both primary and secondary outcrops.
+When the outcrops showed macroscopic differences between the cherts, several samples were collected, in order to obtain as much variability as possible.
 All samples were registered with resource to an android app which allows accurate coordinates to be associated with photos and csv data regarding the outcrop characteristics and conditions (i.e.
-position, abundance, visibility, access, geomorphology, chert morphology and conditions).
-Individual ID tags were associated with each sample and placed in the corresponding bag.
+abundance, visibility, access, geomorphology, chert morphology and conditions).
+All data related to the app and dataset for the variables recorded can be found in the supplementary materials.
+Individual ID tags were associated with each sample, which was washed and labeled in the laboratory.
 
-\*\* outcrops were identified, \*\* in primary and \*\* in secondary deposition.
-From these, \*\* samples were recovered.
-The geological samples were taken to the laboratory where they were washed, relabeled to be inserted in the existing collection and analyzed macroscopically.
+All geological samples were analyzed macroscopically following a ==pre-established dataset==.
+The variables were defined based on specialized literature and the dataset with the variables can be found in the Supplementary materials.
+The goal of this analysis was to describe the cherts and their enclosing rocks in a detailed and objective manner.
+For a comparative analysis with archaeological artefacts, other methods are inconvenient since they are destructive and often difficult to analyze large assemblages.
+Macroscopic analyses have the advantage of being less costly and easy to apply.
+Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops and formations is essential to apply the methodology with archaeological materials and establish comparisons.
+A small hand-lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with the resource of a Nikon SMZ25 stereomicroscope, focusing especially aspects like inclusions and fossil content.
+
+Thin-sections were produced for geological samples of all formations, focusing on obtaining petrographic data which reflected the variability observed at a macroscopic level.
+In total, 29 thin-sections were produced (thin-section database info), divided in three groups: 1) 20 thin-sections of geological samples from different outcrops within the western section of the Algarve, recovered from the 2021 fieldwork; 2) 9 thin-sections of geological samples from different outcrops from the eastern section of the Algarve, recovered from the 2022 fieldwork; 3) 1 thin-section of a geological sample recovered from previous works (\<2021), which was not identified in the latter prospections.
+Although primary outcrops were prioritized, thin-sections of secondary deposition samples were also produced, in order to characterize the chert from most visited locations.
+Information about the thin-sections, including the laboratories in which they were made can be consulted in the Supplementary materials.
+
+All thin-sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in the Supplementary materials).
+Three geological samples from western Algarve (previously broken for thin-section production) were also chosen for X-ray diffraction analysis, which was made at the **CSIC** in Barcelona, using a Bruker D8-A25 with a Cu tube and ultra-fast PSD detector.
+
+All descriptions (macroscopic, petrographic and XRD data) and accompanying photographs will be available online on a database, and downloadable as PDF files.
+The paper was written with resource to R and R Studio and can be found open-source at the **link**.
 
@@
 
+Eighteen outcrops (primary and secondary) were revisited or identified in the Algarve region, nine in the westernmost territory and nine to the east (between Loulé and Tavira).
+From these, 70 samples were recovered and analyzed.
+Some of these samples were recovered as isolated finds or in secondary settings.
+
+On the westernmost part of the Algarve, there are mainly cherts from two different formations: Lower Jurassic and Upper Jurassic.
+The latter can be found in a single known outcrop - Praia da Mareta.
+Lower Jurassic outcrops are more common and have been better studied [@ribeiro_evolucao_2005].
+These outcrops are heterogeneous, showing different characteristics and chert colors.
+
+The Lower Jurassic cherts can be grouped in three main macroscopic types: yellow and/or red type; grey/brown type; yellow and red with fossils type.
+These macroscopic types not only show differences in color, but also in the fossil content, visible at naked eye and stereomicroscope.
+The first two types are present in all outcrops.
+They are mainly characterized by dull to medium luster and opaque translucency, although some samples were sub-translucent.
+The feel ranges between the smooth and semi-smooth, although many of the Belixe cherts have a rough feel.
+In the Yellow/Red cherts, fossil content is present but visible only as white, red or yellow speckling.
+The Grey/Brown cherts show little fossil content, barely visible with the stereomicroscope.
+The Yellow/Red with fossils cherts show a large quantity of big fossils, which are easily seen at naked eye and can be identified under the stereomicroscope.
+
+The Lower Jurassic cherts of Western Algarve are composed mainly by microcrystalline quartz, with textures that range mostly from the wackestone to the packstone.
+In more than 50% of the samples, no fossil can be identified, as all fossils, albeit common to very frequent, are poorly preserved, filled with chalcedony or quartz, and without any identifiable morphology.
+Whenever identifiable, fossils present in the sample are: Echinoderms, Radiolarians, Sponge spicules and a Bivalve shell.
+
+Despite the similar characteristics between these cherts, independent of color or geography, the outcrops are heterogeneous and show varying characteristics, which may be of importance to distinguish between chert sources within the Lower Jurassic formation.
+These outcrops have been divided in four groups, following the available literature: Cabo de S.
+Vicente (including Cabo de S.
+Vicente and Aspa); Foz dos Fornos; Ponta dos Altos; Praia do Belixe (which includes Belixe Sul); and Ferrel.
+
+The Cabo de S.
+Vicente (CSV) and Aspa (ASP) chert is characterized by abundant nodules in the natural rock banks of the cliffs, appearing as horizontal layers within the parent rock.
+The banks seem to be mainly dolomite or dolomitic limestones.
+The process of dolomitization seems to have affected the chert nodules, as they often present different levels of silicification from the edges of the nodule to the interior (which is more silicified), which also affects the knapping quality.
+The nodules vary in size, ranging from small 4 cm in diameter circular nodules to bed-like groups of nodules of \~20 cm width.
+At the Aspa outcrops, the nodules are less frequent and smaller.
+Due to the proximity to the cliffs, the visibility of the chert nodules is good, and in present times, small chunks of chert (without cortex or with small amounts of parent rock attached) accumulate in secondary deposition nearby.
+
+Foz dos Fornos (FZF) and Ponta dos Altos (PdA) show similarities to the CSV outcrops.
+The nodules are visible in several banks of dolomite, dolomitic limestone and limestone, partially covered by soil.
+The nodules can be circular, around 5 cm diameter, or wide with nearly 20 cm of width.
+Despite their size, these cherts are frequently filled with fractures which fragment the larger nodules into smaller volumes of raw material.
+Alike CSV, FZF and PdA also shows cherts with differing degrees of dolomitization, although in apparent smaller quantities than CSV.
+Besides the abundant presence of primary outcrops, there are also abundant chert nodule fragments in secondary deposition, down the slope of the cliff (in the case of FzF) or at the top of the cliff, on a sand path (in the case of PdA).
+These are small, between 1-4 cm of width, but of easy access.
+Between the FZF chert and the PdA, the main differences seem to be the cortex and parent rock, which show differing reactions to hydrocloric acid, the first being a dolomite or dolomitic limestone, and the second being mostly a limestone, with some degree of dolomitization in certain areas.
+
+Praia do Belixe (PBLX) is characterized by the abundance of chert nodules throughout the dolomite layers of the cliff area.
+They are visible in certain areas of the cliff and within the rock shelters.
+The nodules can be small, around 5 cm of diameter, sometimes reaching more than \~30 cm of width, or bedded, as chert layers between the dolomite layers.
+The cherts show varying degrees of dolomitization, and are mostly characterized by coarse to semi-smooth feel, dull luster and a medium to low knapping quality.
+Unlike the other outcrops, no chert nodule fragments were found close to the cliffs, and sample could only be recovered directly from the embedded nodules in the cliff walls.
+Nodules scattered on the floor were only located at Belixe Sul (BLS), a primary outcrop nearly destroyed located on a field, north from the beach area.
+The chert in this outcrop showed no differences from PBLX, aside from the size of the nodules, which were smaller and often showed signs of post-depositional alterations.
+
+A third location for chert has been previously identified north of BLS.
+Belixe Norte (BLN) is located on a dirt road and unused agriculture field.
+Several chert fragments were collected in this location.
+However, BLN is in proximity to an archaeological site and several collected samples were lithic artefacts.
+No larger nodules or outcrop were identified in this location.
+The samples recovered from the location also seem to corroborate that BLN should not be considered an outcrop, as they do not match the local cherts and rather, resemble most the samples recovered from Eastern Algarve.
+
+Ferrel, unlike the other outcrops, is located inland and away from the coast.
+Due to its location in a homonimous village, the state of the outcrop is poor, and all samples were either recovered as scattered nodules or from larger blocks of rock, from a partially destroyed outcrop.
+The proximity of an archaeological site from the -- nearby also raises questions regarding the nodules found in secondary deposition, as these may be surface finds.
+Despite these caveats, the recovered samples are similar to those from the other outcrops, albeit with a better quality, being characterized with a shiny to medium luster and smooth to semi-smooth feel.
+All surface fragments and nodules were small, with around 2 to 3 cm of width which may be explained by the state of the outcrop.
+
+Unlike the outcrops from the Lower Jurassic of western Algarve, this region of the Algarve has only one identified outcrop for Upper Jurassic cherts.
+These are located at Praia da Mareta (MAR) and abundant, or in secondary deposition at Ponta da Atalaia (PtA).
+At Praia da Mareta the nodules are only easily accessible in the beach, where large chunks of the cliff (\~1 m in diameter) are transported by the waves.
+Several chert nodules of different sizes can be found in the parent rock washed ashore, ranging between 2 cm to 20 cm in diameter.
+The quality of the chert also varies, possibly related to different dolomitization stages of the nodules, although this may also be influenced by chemical and physical alterations to the chert.
+At Ponta da Atalaia the chert can be found atop the cliffs, with rare nodules scattered on the floor.
+The Upper Jurassic cherts are very similar to the Lower Jurassic, with dull to medium luster and grey/purple colors.
+The translucency ranges from opaque to areas where the chert is translucent.
+This translucency may be a significant difference do distinguish between outcrops.
+Petrographically, the cherts are also similar to the Lower Jurassic ones.
+The only identifiable difference is the presence of calcispheres.
+All samples from the Mareta outcrops seen under the petrographic microscope showed the presence of abundant calcispheres, which is not always apparent with the stereomicroscope.
+Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic, which were uncommon and scattered at the top of the cliffs by the beach.
+
+On the eastern part of the Algarve, chert-bearing known formations are from the Middle to Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively.
+
+The Malhão formation chert (dating to the Middle Jurassic) was identified in three outcrops, in the Faro parsish and the Tavira parish.
+Whenever in a primary outcrop, this chert was homogeneous.
+The secondary deposits were recent waterlines and slope deposits, and the cherts were often characterized by intense post-depositional alterations.
+In these cases, it was not possible to confirm the outcrop location.
+In these outcrops, the nodule frequency varied from common to abundant.
+The nodules are roundish, ranging between 3 to 5 cm of maximum width.
+In all cases, access to the outcrops was easy.
+Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits.
+The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts.
+In general, they are both characterized by a dull to medium luster, Opaque to Sub-translucent translucency and smooth to semi-smooth feel.
+They are easily identifiable through the high amounts of macroscopically visible inclusions, which look like white Speckling in plain sight.
+Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified.
+The petrographic analysis shows for the Malhão cherts from Casal da Colina (primary and secondary outcrops) high amounts of dolomite within the chert.
+All of these cherts are characterized by a wackestone texture and high variety of identifiable fossils (although all poorly preserved and replaced by chalcedony or quartz).
+These fossils are: Sponge spicules, Radiolarians, Ostracods, Tentaculites (also previously identified by --), Echinoderms and Calcispheres.
+
+The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve, above Moncarapacho in the Olhão parish.
+Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock.
+No chert was identified in any secondary deposits, which might be related to the anthropic alteration of the landscape.
+Smaller nodules broken from the parent rock were identified near the primary source in a field.
+Whenever embedded in the parent rock, the nodules varied in size (\~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock.
+The cherts show little macroscopic variability between nodule and outcrop.
+They are grey/brown (with visible yellow inclusions).
+Within nodule however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas.
+Some of the nodules also show a variability of translucency, with areas which are translucent, with a very fine grain and little presence of visible inclusions.
+The petrographic analysis shows that the cherts range from a wackestone to packstone texture, which was already seen macroscopically.
+They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals.
+Fossils are poorly preserved in general, with a few being identifiable: calcispheres, bivalve shell, sponge spicules, Ostracod, Echinoderms and gastropod.
+
 # Discussion
+
+The prospection works and analyses to the collected geological samples show that the south of Portugal has high potential for raw material studies, especially regarding chert.
+The presence of chert-bearing outcrops in the westernmost part of the Algarve, center and east would provide several possibilities for sourcing and procurement whenever groups moved throughout the territory.
+This is further important when we consider the geology of this territory.The geology of the Algarve itself may have played an important part in how groups procured their raw materials, specifically, their chert, a task which has been identified as essential for hunter-gatherer groups.
+To the south, communities would only have access to chert-bearing outcrops up to the coast.
+To the north, the mountain range would not only have provided no chert nodules, but may have also hampered the movement of populations, forcing groups to move east and west instead of north or south.
+This movement may have facilitated the gathering of cherts from different formations within the Algarve, posteriorly then brought into the sites.
+Especially for Middle and Upper Paleolithic occupations, understanding the sources of chert in the Algarve may provide data about where in the territory they were sourcing their raw materials, and how they were using the territory when their movement was constrained by the Peninsula's natural barriers.
+This question remains unanswered, but raw material data and comparative studies between archaeological assemblages and geological samples may provide the necessary results to answer it.
+
+However, tracking these movements and procurement patterns would only be possible if the cherts from the different formations and outcrops could be traced back to their sources.
+This presented itself as a first caveat for this type of studies, since for the Algarve, for example, all cherts come from Jurassic formations in pelagic environments.
+Despite the similar formation environments, n general there seems to be relevant differences between the cherts of different formations and periods.
+This is further relevant given the fact that they are geographically distant.
+Within formations, however, there are no discernable differences, both at a macroscopic and petrographic level, as these do not seem to be useful to distinguish between outcrops.
+That is most obvious on the Lower Jurassic outcrops of west Algarve.
+The identified chert groups, which varied mostly in color and fossil content, are present in several outcrops.
+In this region, the variables which may be better used to understand which outcrops were visited may be the quality and size of the nodules.
+The latter, for example, is an important variable in the Belixe outcrops, which show the largest volumes of rock, even if the knapping quality is worse than some other available, smaller nodules.
+Size may be used in conjunction with other technological data, to understand whether different nodules were being explored differently based on their size, or their procurement was being preferred in relation to other smaller nodules in possibly closer outcrops in the region.
+
+The Upper Jurassic nodules of western Algarve also show larger volumes than those from the Lower Jurassic.
+This, together with the translucency, which is often characterized by sub-translucent fabrics, may be a good macroscopic indicator for the chert's provenance.
+However, for a reliable distinction between the Lower Jurassic and the Upper Jurassic cherts, petrographic analyses and the identification of calcispheres may be necessary.
+
+The differences identified between the cherts of the various formations can be seen both at a macroscopic and petrographic level.
+Given the formation setting, petrographically, all the cherts from the Algarve are fairly homogeneous - marine origin, in limestone or dolomitic limestone rocks, all formed during the Jurassic.
+The use of fossils for the identification of the cherts is also difficult, since these are often not well preserved enough to allow the identification of species that may connect a group of cherts.
+The size, frequency and preservation state of the fossils seems to be, then, one of the defining criteria for discerning cherts from different formations, and thus, different geographic areas.
+These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana and Malhão - to be differentiated without the need for thin-sections.
+This is specially important for archaeological collections, specially those which may be small, with small archaeological pieces, or for the study of older collection to which other means of analysis may not be available.
+
+This data seems to confirm the potential of a macroscopic analysis to study the cherts of the Algarve.
+Albeit applying different methodologies, such as petrographic analyses, to these cherts is a way of completing the petrographic study of a collection, applying reliably mostly a macroscopic analysis to the assemblages coming from southern Portugal helps tackle issues such as the destructiveness, costliness and time consumption of some methods.
+
+The research was able to complete the collection, providing samples of the available cherts which may be compared with archaeological samples.
+However, there are some caveats to this type of comparative studies.
+Landscapes have changed through time, specially with the influence of modern societies.
+House constructions, agricultural fields and roads, for example, have modified the landscape, possibly altering the availability and visibility of raw materials.
+The existence of chert in specific points in the map may not necessarily mean that they do not exist elsewhere.
+However, identifying characteristics which are common to most, if not all cherts of a region, may be a good indication that some cherts are not local.
+
+Another caveat regarding chert sources, and specially in a geographic area like the Algarve, is the possibility of some outcrops being submerged.
+Previous studies have identified the existence of Jurassic outcrops with chert nodules on the west coast, submerged by water.
+In times where the sea level was similar to the current one, these outcrops would not have been accessible, even during low tide, the only possibility for their procurement in nodules dragged to the shore.
+However, during periods where the sea level was lower due to water freezing in the polar caps (during the LGM, for example), large portions of the coast would have been accessible.
+This raises the question whether the chert reference collection is actually representative of all the variability of chert in the region, including the submerged coastal areas.
+For studies focusing periods like the Proto-solutrean and Solutrean in the Algarve, for example, where nearly 20 km of shore was exposed at the peak of the LGM, there may have been the exploitation of unknown sources or cherts with more variability than that known from the recovered samples.
+Despite the caveat, this raises the possibility to understand whether this new portion of landmass altered the raw material procurement patterns of these groups, or added new resources which had been previously unavailable.
+Studies which compare Gravettian and Magdalenian assemblages to Proto-solutrean and Solutrean assemblages, within one single site, may give new insights to this question.
 

### 2022-09-20 analysis/supplementary-materials/PetrologicalAnalysis.Rmd
- First draft of article text without corrections

---
title: "Thin section analysis"
author: "Joana Belmiro"
date: "5/20/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r library, include=FALSE}
library(knitr)
library(dplyr)
library(readr)
library(kableExtra)
library(tidyr)
```

```{r readbd, include=FALSE}

TSdb <- read_csv(here::here('analysis/data/raw_data/PetrographyDB.csv')) # RM types file

```

# Geological samples

## SP6

Sample `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP6") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP6") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP6") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ORTHOCHEM2TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP6") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP6") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP6") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP6") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP6") %>% select(BIOCLAST2))`. There is `r as.character(TSdb %>% filter(ID == "SP6") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP6") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP6") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP7

Sample `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP7") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP7") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP7") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ORTHOCHEM3TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP7") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP7") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP7") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP7") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP7") %>% select(BIOCLAST2))`. There is `r as.character(TSdb %>% filter(ID == "SP7") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP7") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP7") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP9

Sample `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP9") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP9") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP9") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ORTHOCHEM2TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP9") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP9") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP9") %>% select(BIOCLAST1))`. There is `r as.character(TSdb %>% filter(ID == "SP9") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP9") %>% select(POROSITYTYPE))` type. Other sedimentary features are a `r as.character(TSdb %>% filter(ID == "SP9") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP10

Sample `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP10") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP10") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP10") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ORTHOCHEM2TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ALLOCHEM1))`; `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ALLOCHEM2FREQ))` `r as.character(TSdb %>% filter(ID == "SP10") %>% select(ALLOCHEM2))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP10") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP10") %>% select(BIOCLAST1))`. There is `r as.character(TSdb %>% filter(ID == "SP10") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP10") %>% select(POROSITYTYPE))` type. Other sedimentary features are a `r as.character(TSdb %>% filter(ID == "SP10") %>% select(SEDIMENTARYSTRUCTURES))`. `r as.character(TSdb %>% filter(ID == "SP10") %>% select(OBSERVATIONS))`.

## SP14

Sample `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP14") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP14") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP14") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ORTHOCHEM3TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP14") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP14") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP14") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP14") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP14") %>% select(BIOCLAST2))`; `r as.character(TSdb %>% filter(ID == "SP14") %>% select(BIOCLAST3FREQ))` `r as.character(TSdb %>% filter(ID == "SP14") %>% select(BIOCLAST3))`. `r as.character(TSdb %>% filter(ID == "SP14") %>% select(OBSERVATIONS))`. There is `r as.character(TSdb %>% filter(ID == "SP14") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP14") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP14") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP18

Sample `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP18") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP18") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP18") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM3TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM4FREQ))` of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM4))`, of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ORTHOCHEM4TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ALLOCHEM1))`; `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ALLOCHEM2FREQ))` `r as.character(TSdb %>% filter(ID == "SP18") %>% select(ALLOCHEM2))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP18") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP18") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP18") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP18") %>% select(BIOCLAST2))`; `r as.character(TSdb %>% filter(ID == "SP18") %>% select(BIOCLAST3FREQ))` `r as.character(TSdb %>% filter(ID == "SP18") %>% select(BIOCLAST3))`. There is `r as.character(TSdb %>% filter(ID == "SP18") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP18") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP18") %>% select(SEDIMENTARYSTRUCTURES))`. `r as.character(TSdb %>% filter(ID == "SP18") %>% select(OBSERVATIONS))`.

## SP21

Sample `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP21") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP21") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP21") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM3TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM4FREQ))` of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM4))`, of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ORTHOCHEM4TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP21") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP21") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP21") %>% select(BIOCLAST1))`. `r as.character(TSdb %>% filter(ID == "SP21") %>% select(OBSERVATIONS))`. There is `r as.character(TSdb %>% filter(ID == "SP21") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP21") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP21") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP23

Sample `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP23") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP23") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP23") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM3TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM4FREQ))` of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM4))`, of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ORTHOCHEM4TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP23") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST2))`; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST3FREQ))` `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST3))`; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST4FREQ))` `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST4))`; `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST5FREQ))` `r as.character(TSdb %>% filter(ID == "SP23") %>% select(BIOCLAST5))`. `r as.character(TSdb %>% filter(ID == "SP23") %>% select(OBSERVATIONS))`. There is `r as.character(TSdb %>% filter(ID == "SP23") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP23") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP23") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP27

Sample `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP27") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP27") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP27") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ORTHOCHEM3TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP27") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP27") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP27") %>% select(BIOCLAST1))`.  `r as.character(TSdb %>% filter(ID == "SP27") %>% select(OBSERVATIONS))`. There is `r as.character(TSdb %>% filter(ID == "SP27") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP27") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP27") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP32

Sample `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP32") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP32") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP32") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ORTHOCHEM2TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ALLOCHEM1))`; `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ALLOCHEM2FREQ))` `r as.character(TSdb %>% filter(ID == "SP32") %>% select(ALLOCHEM2))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP32") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP32") %>% select(BIOCLAST1))`. `r as.character(TSdb %>% filter(ID == "SP32") %>% select(OBSERVATIONS))`. There is `r as.character(TSdb %>% filter(ID == "SP32") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP32") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP32") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP34

Sample `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP34") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP34") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP34") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM3TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM4FREQ))` of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM4))`, of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ORTHOCHEM4TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP34") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST2))`; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST3FREQ))` `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST3))`; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST4FREQ))` `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST4))`; `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST5FREQ))` `r as.character(TSdb %>% filter(ID == "SP34") %>% select(BIOCLAST5))`. `r as.character(TSdb %>% filter(ID == "SP34") %>% select(OBSERVATIONS))`. There is `r as.character(TSdb %>% filter(ID == "SP34") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP34") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP34") %>% select(SEDIMENTARYSTRUCTURES))`.

## SP39

Sample `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP39") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP39") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP39") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ORTHOCHEM3TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ALLOCHEM1))`; `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ALLOCHEM2FREQ))` `r as.character(TSdb %>% filter(ID == "SP39") %>% select(ALLOCHEM2))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST2))`; `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST3FREQ))` `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST3))`; `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST4FREQ))` `r as.character(TSdb %>% filter(ID == "SP39") %>% select(BIOCLAST4))`. There is `r as.character(TSdb %>% filter(ID == "SP39") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP39") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP39") %>% select(SEDIMENTARYSTRUCTURES))`. `r as.character(TSdb %>% filter(ID == "SP39") %>% select(OBSERVATIONS))`.

## SP40

Sample `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP40") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP40") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP40") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM2TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM3FREQ))` of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM3))`, of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ORTHOCHEM3TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ALLOCHEM1))`; `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ALLOCHEM2FREQ))` `r as.character(TSdb %>% filter(ID == "SP40") %>% select(ALLOCHEM2))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST2))`; `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST3FREQ))` `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST3))`; `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST4FREQ))` `r as.character(TSdb %>% filter(ID == "SP40") %>% select(BIOCLAST4))`. There is `r as.character(TSdb %>% filter(ID == "SP40") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP40") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP40") %>% select(SEDIMENTARYSTRUCTURES))`. `r as.character(TSdb %>% filter(ID == "SP40") %>% select(OBSERVATIONS))`.

## SP42

Sample `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ID))` is a `r as.character(TSdb %>% filter(ID == "SP42") %>% select(LITHOLOGY))`, characterized by a `r as.character(TSdb %>% filter(ID == "SP42") %>% select(TEXTURE))` texture and a `r as.character(TSdb %>% filter(ID == "SP42") %>% select(MICROSTRUCTURE))` microstructure. `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEMCOUNT))` orthochems were identified: `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEM1FREQ))` of `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEM1))`, of `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEM1TYPE))` type; `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEM2FREQ))` of `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEM2))`, of `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ORTHOCHEM2TYPE))` type. Allochems are `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ALLOCHEM1FREQ))` `r as.character(TSdb %>% filter(ID == "SP42") %>% select(ALLOCHEM1))`. Bioclasts are `r as.character(TSdb %>% filter(ID == "SP42") %>% select(BIOCLAST1FREQ))` `r as.character(TSdb %>% filter(ID == "SP42") %>% select(BIOCLAST1))`; `r as.character(TSdb %>% filter(ID == "SP42") %>% select(BIOCLAST2FREQ))` `r as.character(TSdb %>% filter(ID == "SP42") %>% select(BIOCLAST2))`. There is `r as.character(TSdb %>% filter(ID == "SP42") %>% select(TOTALPOROSITY))` of total porosity of `r as.character(TSdb %>% filter(ID == "SP42") %>% select(POROSITYTYPE))` type. Other sedimentary features are `r as.character(TSdb %>% filter(ID == "SP42") %>% select(SEDIMENTARYSTRUCTURES))`.

### 2022-09-20 analysis/paper/paper.Rmd
- First corrections XT

--- 2022-05-18 analysis/paper/paper.Rmd
@@
 ---
-title: "Title Goes Here"
+title: "Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia."
 author:
-  - Myfirstname Mylastname:
+  - name: Joana Belmiro
     email: fl@oneplace.edu
@@
 
+```{r packages, include=FALSE}
+
+library(ggplot2)
+theme_set(theme_bw())
+library(sf)
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(readr)
+library(ggspatial)
+library(ggrepel)
+library(dplyr)
+library(kableExtra)
+library(flextable)
+library(data.table)
+library(tidyr)
+
+```
+
+```{r databases, include=FALSE}
+
+petrography <- read_csv(here::here('analysis/data/raw_data/PetrographyDB.csv')) # Reading the petrography data
+chertfield <- read_csv(here::here('analysis/data/raw_data/chert_field.csv')) # Reading the field data
+
+fielddata <- chertfield %>% # Cleaning the Field data
+  mutate(Geography = case_when(long <= -8.2 ~ "Ocidental",
+                               long > -8.2 ~ "Oriental")) %>% 
+  filter(`Raw Material` == "Chert") %>% 
+  rename(SampleID = NewCode)
+
+```
+
+```{r map setup, include=FALSE}
+
+sf::sf_use_s2(FALSE)
+world <- ne_countries(scale = "medium", returnclass = "sf")
+#europe <- world[world$continent == "Europe",]
+europe_cropped <- st_crop(world, xmin = -15, xmax = -7.5,
+                                    ymin = 35, ymax = 37.6)
+
+```
+
+```{r map, echo=FALSE, include=FALSE}
+
+ggplot(data = europe_cropped) +
+  geom_sf() +
+  geom_point(data = chertfield, aes(x = long, y = lat, shape = State, color = State), size = 4) +
+  #geom_label_repel(data = chertsamples, aes(long, lat, label=NewCode), box.padding = 0.5, max.overlaps = 30) +
+  annotation_scale(location = "br", width_hint = 0.5) +
+  annotation_north_arrow(location = "br", which_north = "true", 
+        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
+        style = north_arrow_fancy_orienteering)
+  #coord_sf(xlim = c(long_min, long_max), ylim = c(lat_min,lat_max), expand = FALSE)
+
+```
+
+```{r databasesetup, include=FALSE}
+
+CStable <- chertfield %>% 
+  select(NewCode, Site, `Sample variability`, State, Deposit, Type, Frequence, `State of site`, `Thin-sections`)
+
+```
+
+```{r cstable, echo=FALSE, include=FALSE}
+
+regulartable(CStable) %>% 
+  theme_zebra()
+
+```
+
 # Introduction
 
-Here is a citation [@Marwick2017]
+The South Iberian territory is a key area to understand cultural transitions from the Middle Paleolithic to early Prehistory .
+The Algarve region specifically (south Portugal) has contributed greatly to understanding pre-history at a European level, with several well-known archaeological sites and publications.
+As a territory located at the tip of the European continent and with a generally warm climate even during the coldest periods which characterize the Late Pleistocene, it has been regarded as one of the most significant glacial refugia in Europe (Hewitt, 2000; Gómez and Lunt, 2007; González-Sampériz et al.
+2010).
+Several authors have suggested that the Middle to Upper Paleolithic cultural sequences present in the Iberian Peninsula territory are related to adaptation and reorganization processes related to abrupt climatic events (Brädtmoller et al., 2012; Cascalheira et al., 2021).
+As such, the south Iberian territory witnessed several moments of crisis, cultural changes, and behavioral adaptations.
+An example of this is the late occupation of the territory by Neanderthal populations and their consequent disappearance.
+The Iberian record for the Middle Paleolithic allows for nearly 250 kyr of Neanderthal occupations to be studied until the extinction interval (de La Torre, Martínez-Moreno, and Mora, 2013).
+Several studies suggest that Neanderthal presence in southern Iberia lasted until c.
+37 ka cal BP or later, with a different trajectory than that found in other places of Europe (Cascalheira, Gonçalves, and Maio, 2021).
 
-## Previous research
+The several cultural shifts of the Upper Paleolithic (UP) are also represented in the region, in sites with long chronological occupations, from the first UP occupations such as the Aurignacian (not detected in southern Portugal) and the Gravettian, such as the archaeological site of Vale Boi .
+This site, containing one of the most complete UP sequences of the Iberian Peninsula, offers the opportunity to understand the movements of the first groups of Humans to reach the territory, their cultural changes throughout the Upper Paleolithic, as well as the impact of abrupt climatic change in culture, such as the Heinrich Event 2 (HE 2) and the Last Glacial Maximum (LGM).
+Finally, south Iberia also shows the existence of several Mesolithic, Neolithic, and Chalcolithic settlements and occupations.
+In south Portugal, for example, studies have shown the use of riverine and estuarine areas for Mesolithic settlements (Carvalho, 2008) and a concentration of sites by the westernmost coast of the Algarve (Bicho, 2009), as well as clear settlement and cultural shifts in the Neolithic occupations, followed by Chalcolithic occupations (Sánchez et al., 2012).
+This complete recent prehistory chronology makes southern Iberia, and in specific, southern Portugal, an ideal territory to study the replacement of the last hunter-gatherers of Europe by the first farmers, and the subsequent transformations these communities went through.
 
-## Geological setting and chert outcrops
+From at least ~40 000 cal BP to 3 000 cal BP (** , Valera, 2012), South Portugal has seen moments of crisis and cultural change of different kinds, which gives this territory a large potential for prehistorical archaeology studies and the study of human behavior in a wide chronological range.
+With such a wide chronology, differences in the exploitation of resources, management of territories, settlement patterns, subsistence techniques, climate, technology, and culture, it is important, however, to find a unifying subject that can be studied, analyzed, and compared, bringing together several studies and sites to create a more unified story of the evolution of human behavior.
+That subject may be the raw material that was used by Neanderthals, Human hunter-gatherers, and farmers - that raw material is chert.
+Whenever geologically present (and at times, even when it is not geologically present), Chert has served as a raw material for the production of blanks to more specialized and complex tools.
+As a good quality rock, it was procured from close to far-away locations, transported, knapped, retouched, re-used, heated, and sometimes traded, traveling long distances in the hands of different groups which produced them.
+Countless archaeologists and studies have shown the potential of this raw material to understand cultural aspects such as foraging strategies and territory use, mobility, shifts in technological necessities, cultural preferences, changes in raw material visibility and accessibility, or even the establishment, expansion, and shrinkage of social networks ** .
+As a raw material, chert has been studied in South Portugal.
+The study of chert-bearing outcrops has been a topic of interest for both geologists and archaeologists in the Algarve region, albeit for different reasons.
+Geological literature has often focused on the processes that formed the seas and landmasses of the Algarve basins, as well as the paleobiology and paleobiogeography of the region.
+Archaeological literature has focused on chert sources as a way to predict the availability of mineral resources, and the existence of settlements, characterize hunter-gatherer culture throughout prehistory and understand group mobility and social exchange through time.
+Geological studies in the Algarve region go back to the 19th century and throughout all of the 20th century, through the creation of geological cartography.
+These first works provided the basis for the geological knowledge of the Algarve basin, and it allowed further studies within the geosciences, as a large number of Doctoral theses.
 
-### Geological setting
+Unlike what happens in other areas of Portugal (add references from Aubry, Abrunhosa, etc), archaeological studies focusing on raw materials, especially chert, have been more scarce.
+In general, lithic technology studies in the Algarve have approached raw materials through a macroscopic methodology, often related to the location of the settlement (i.e., [@bicho_o_2003]; Pereira et al 2016 ) or the description of the collection (i.e., [@cascalheira_tecnologia_2010], [@marreiros_lithic_2015]).
+The work of Veríssimo (2005) focusing on the occurrence of chert in western Algarve provided the initial basis for comparative studies, maintaining however solely a macroscopic component.
+Finally, the creation of LusoLit (Telmo et al.
+2016) and the collection of samples from the Algarve provided a new leap in the study of chert in the region.
+Given the existence of several chert-bearing outcrops in South Portugal, preliminary raw material studies, a region of archaeological importance, and the existence of a rich archaeological record, the Algarve stands as a region full of potential for studies of chert and the development of a lithotheque which may serve for regional studies, but also for comparative studies outside, where chert may have been transported and traded, offering the chance to complete the record of exotic cherts which may be coming from the Algarve.
 
-The Algarve region is located in south Portugal, framed north and east by the Alentejo region and Spain respectively.
+This paper aims to locate and characterize the nature of chert sources in southern Portugal and their availability.
+Furthermore, it aims to test the potential of different methodological approaches to characterize the regional cherts and develop a methodological approach adapted to this study area.
+
+# Geological setting and chert outcrops
+
+## Geological setting
+
+The Algarve region is located in southern Portugal, framed north and east by the Alentejo region and Spain respectively.
 To the west and south, it is bordered by the Atlantic ocean.
 Geologically, it is composed of two main geological units: the South Portuguese Zone (SPZ) and the Algarve basin.
-The SPZ is composed of four "domains" of different lithologies and ages, extending up to Alentejo, although all within the Paleozoic period.
-It is overlain unconformably by the Mesozoic sedimentary rocks of the basin [\@fernandes_new_2012].
-The Algarve basin is one of broader Jurassic outcrops in Portugal [@malchus_microfacies_1988].
-It corresponds to the mesozoic-cenozoic sediments which outcrop south of Portugal, from Cape of St.
+The SPZ is composed of four “domains” of different lithologies and ages, extending up to Alentejo, although all within the Paleozoic period.
+It is overlain unconformably by the Mesozoic sedimentary rocks of the basin [@fernandes_new_2012].
+The Algarve basin is one of the broader Jurassic outcrops in Portugal (Malchus and Kuss, 1988).
+It corresponds to the Mesozoic-cenozoic sediments that outcrop south of Portugal, from Cape of St.
 Vicent to Vila Real de St.
 António.
-The Algarve basin is associated with the opening of the central Atlantic and with the eventual oceanic crust formation in the western part of the Tethis sea, between the Algarve and North Africa [@terrinha_bacia_2013].
-After the sedimentation period between the Triassic and the Sinemurian (Lower Jurassic), the mesozoic basin was divided in two sub-basins (western sub-basin and oriental sub-basin), until the Calovian (Middle Jurassic), when an uplift episode allowed a uniform sedimentation of the basin [@rocha_estudo_1976; @terrinha_bacia_2013].
-The existence of two sub-basins and a structural higher area between them, allied with the expansion and retraction of the seas allowed a variation of sedimentation environments, ranging between external platform and internal platform, and continental, hemipelagic, or deep marine environments [@terrinha_bacia_2013].
+The Algarve basin is associated with the opening of the central Atlantic and with the eventual oceanic crust formation in the western part of the Tethys sea, between the Algarve and North Africa (Terrinha et al., 2013).
+After the sedimentation period between the Triassic and the Sinemurian (Lower Jurassic), the Mesozoic basin was divided into two sub-basins (western sub-basin and oriental sub-basin).
+The existence of two sub-basins and a structural higher area between them, allied with the expansion and retraction of the seas allowed a variation of sedimentation environments, ranging between the external platform and internal platform, and continental, hemipelagic, or deep marine environments (Terrinha et al., 2013).
 It was this variance in deposition environments that created the variety of Paleozoic sedimentary facies, with moments of more or less homogeneity throughout this period.
+The Algarve basin is of essential importance for the existence of chert in the Algarve region since it is in the Paleozoic sediments where chert nodules can be found.
+Paleozoic sedimentation in the basin starts in the Triassic, although the identification of chert nodules or beds is limited to the Lower to Upper Jurassic.
+It is at the start of the Lower Jurassic, during the Lower Pliesbachian (also named Carixian) that the Algarve basin started to separate into two sub-basins until the Callovian (Middle Jurassic) when an uplift episode allowed uniform sedimentation of the basin (Rocha, 1976; Terrinha et al., 2013).
+During the Lower Pliensbachian, the sediments in the western sub-basin can be described as marine, of the external platform, while the sediments of the eastern sub-basin are of an internal platform.
+The transition between the Lower Jurassic to the Middle Jurassic shows a hiatus, as there are no known sediments from the Toarcian to the Aaelinian in the basin, the sedimentation restarting only after the latter age.
+During the Upper Jurassic, the lithofacies variation became more prominent, especially during the Upper Oxfordian-Lower Kimmeridgian, with a certain uniformity of the lithofacies in all the sub-basins during the Upper Kimmeridgian (Terrinha et al., 2013).
 
-The Algarve basin is of essential importance for the existence of chert in the Algarve region, since it is in the Paleozoic sediments where chert nodules can be found.
-Paleozoic sedimentation in the basin starts in the Triassic, although the identification of chert nodules or beds is limited to the Lower to Upper Jurassic.
-It is at the start of the Lower Jurassic, during the Lower Pliesbachian (also named Carixian) that the Algarve basin started to separate into two sub-basins, until the Calovian (Middle Jurassic), when an uplift episode allowed a uniform sedimentation of the basin [@rocha_estudo_1976; @terrinha_bacia_2013].
-During the Lower Pliesbachian, the sediments in the western sub-basin can be described as marine, of external platform, while the sediments of the eastern sub-basin are of an internal platform.
-The transition between the Lower Jurassic to the Middle Jurassic shows a hiatus, as there are no known sediments from the Toarcian to the Aaelinian in the basin, the sedimentation restarting only after the latter age.
-During the Upper Jurassic the lithofacies variation became more prominent, especially during the Upper Oxfordian-Lower Kimmeridgian, with a certain uniformity of the lithofacies in all the sub-basins during the Upper Kimmeridgian [@terrinha_bacia_2013].
-
-### Chert outcrops
+## Chert outcrops
 
@@
 They appear with relative frequency, and often as nodules of small to medium dimensions, although thin beds of chert can also be identified in some areas.
-
 As mentioned above, outcrops with chert nodules in the Algarve region can be found in the Paleozoic sediments from the Algarve Basin, more specifically in the Jurassic sediments.
@@
 
-Lower Jurassic geological layers with chert nodules can be mostly found in the Carixian-age sediments (also known as Pliensbachian) from the Sagres region, often outcropping in the cliffs by the beaches.
+Lower Jurassic geological layers with chert nodules can be mostly found in the Carixian-age sediments (also known as Pliensbachian) from the Sagres region, often outcropping in those places where more layers are exposed, such as beach-generated cliffs and associated deposits.
 These include the chert nodules and beds from Cabo de S.
 Vicente to Praia do Belixe, but also the small outcrop inland, in Ferrel.
-The chert nodules occur within limestone, dolomitic limestone and marly limestone.
-Further east one layer with chert nodules was also identified, but attributed to the Sinemurian-Toarcian, in a formation identified as "Limestones with *Paleodasucladus* and limestones with chert nodules".
-Described only in the Geologic Map at a 1 200 000 scale however, these chert nodules were characterized as micronodules [\@oliveira_carta_1992].
+The chert nodules occur within the limestone, dolomitic limestone, and marly limestone.
+Further east one layer with chert nodules was also identified but attributed to the Sinemurian-Toarcian, in a formation identified as “Limestones with Paleodasucladus and limestones with chert nodules”.
+Described only in the Geologic Map at a 1 200 000 scale, however, these chert nodules were characterized as micronodules [@oliveira_carta_1992].
 
-Middle Jurassic geological layers with chert nodules are only found east of Sagres, in the Albufeira, Faro and Tavira region.
+Middle Jurassic geological layers with chert nodules are only found east of Sagres, in the Albufeira, Faro, and Tavira regions.
 These layers were attributed to the Malhão Formation, from the Aalenian-Bajocian age.
-In general, this formation was described as carbonated, marine and thick, and chert can be found in two different layers: monogenic conglomerates with micritic limestone intercalations with chert beds and nodules; microcrystalline limestones with chert nodules (pel-biomicrite).
-Another chert outcrop was identified in the Geologic Map at a 1 200 000 scale, attributed to the Bajocian-Batonian age and named "Alagoa conglomerates", with partially dolomitic limestones and chert nodules.
+In general, this formation was described as carbonated, marine, and thick, and chert can be found in two different layers: monogenic conglomerates with micritic limestone intercalations with chert beds and nodules; microcrystalline limestones with chert nodules (pel-biomicrite).
+Another chert outcrop was identified in the Geologic Map at a 1 200 000 scale, attributed to the Bajocian-Batonian age and named “Alagoa conglomerates”, with partially dolomitic limestones and chert nodules.
 
-Finally, Upper Jurassic sediments have been identified also in the Albufeira, Faro and Tavira area.
-The Jordana Formation was attributed to the Kimmeridgian/Kimmeridgian-Portlandian, and is characterized by dark-gray limestones, with frequent secondary silifications with abundant fossil fragments.
-In the Tavira area these layers were instead attributed to the Oxfordian-Kimmeridgian age.
+Finally, Upper Jurassic sediments have been identified also in the Albufeira, Faro, and Tavira area.
+The Jordana Formation was attributed to the Kimmeridgian/Kimmeridgian-Portlandian and is characterized by dark-gray limestones, with frequent secondary silifications with abundant fossil fragments.
+In the Tavira area, these layers were instead attributed to the Oxfordian-Kimmeridgian age.
 
@@
 
-## Fieldwork and geological samples
-
-To obtain geological samples several known chert outcrops were visited during August of 2021 and April of 2022, to prospect the Southwest and Southeast portion of the Algarve region (respectively).
-The first step of fieldwork included the review of previously known research, including geological maps, archaeological prospection works and raw material studies.
+To identify the chert outcrops present in southern Portugal and understand the chert’s characteristics, a macroscopic and petrographic approach was applied to the study of geological samples.
+A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials.
+This is because different methods have their inherent limitations.
+By combining these two different methodologies the goal was to create several layers of analysis, which may then be further pursued or added new layers (Brandl, 2016).
+Several other similar methodologies and approaches have been applied in other regions (references).
+However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, problematics, and the characteristics of the types of cherts in question.
+Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region, the two methodologies were chosen for the study.
+Geological samples were obtained through fieldwork, from August of 2021 (west Algarve) and April to June 2022 (east Algarve), and included visits to previously known outcrops and prospection of areas identified in geological maps.
+The first step of fieldwork included the review of previously known research, including geological maps, archaeological prospection works, and raw material studies (Pereira, Verissimo 2004, 2005).
 These were followed as a reference guide for recovering samples.
-In the field, several chert nodules were collected within one single outcrop, in order to obtain as much variability as possible.
-The surrounding areas to the outcrops were also surveyed to understand other possible outcrops or secondary deposition environments.
-All samples were registered with resource to an android app which allows accurate coordinates to be associated with photos and csv data regarding the outcrop characteristics and conditions (i.e.
-position, abundance, visibility, access, geomorphology, chert morphology and conditions).
-Individual ID tags were associated with each sample and placed in the corresponding bag.
-
-\*\* outcrops were identified, \*\* in primary and \*\* in secondary deposition.
-From these, \*\* samples were recovered.
-The geological samples were taken to the laboratory where they were washed, relabeled to be inserted in the existing collection and analyzed macroscopically.
+Whenever necessary, samples from previous prospection works were consulted, from the LusoLit lithoteque from ICArEHB at the University of Algarve (Pereira) and the thin sections at the University of Évora.
+Whenever coordinates or specific locations for known outcrops were available, those were visited, and then the surrounding area was prospected to understand the extension of the outcrops and possible secondary deposition outcrops nearby.
+In all other situations, several locations with the potential to find outcrops were selected from polygons where formations with chert were known to the outcrop.
+Samples were collected whenever possible, focusing on both primary and secondary outcrops.
+When the outcrops showed macroscopic differences between the cherts, several samples were collected, to obtain as much variability as possible.
+All samples were georeferenced and registered with resources to an android app which allows accurate coordinates to be associated with photos and CSV data regarding the outcrop characteristics and conditions (i.e.
+abundance, visibility, access, geomorphology, chert morphology, and conditions).
+All data related to the app and dataset for the variables recorded can be found in the supplementary materials.
+Individual ID tags were associated with each sample, which was washed and labeled in the laboratory.
+All geological samples  were analyzed macroscopically following a pre-established dataset.
+The variables were defined based on specialized literature and the dataset with the variables can be found in the Supplementary materials.
+The goal of this analysis was to describe the cherts and their enclosing rocks in a detailed and objective manner.
+For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages.
+Macroscopic analyses have the advantage of being less costly and easy to apply.
+Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops, and formations is essential to apply the methodology with archaeological materials and establish comparisons (Bustillo et al 2009 ).
+A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with the resource of a Nikon SMZ25 stereomicroscope, focusing especially on aspects like inclusions and fossil content.
+Thin sections were produced for geological samples of all formations, focusing on obtaining petrographic data which reflected the variability observed at a macroscopic level.
+In total, 29 thin sections  were produced (thin-section database info), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve, recovered from the 2021 fieldwork; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve, recovered from the 2022 fieldwork; 3) 1 thin-section of a geological sample recovered from previous works (<2021), which was not identified in the latter prospections.
+Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced, to characterize the chert from the most visited locations.
+Information about the thin sections, including the laboratories in which they were made, can be consulted in the Supplementary materials.
+All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in the Supplementary materials).
+Three geological samples from western Algarve (previously broken for thin-section production) were also chosen for X-ray diffraction analysis, which was made at the GeoSciences CSIC institute in Barcelona, using a Bruker D8-A25 with a Cu tube and ultra-fast PSD detector.
+All descriptions (macroscopic, petrographic, and XRD data) and accompanying photographs will be available online on a database, and downloadable as PDF files.
+The paper was written with resource to R and R Studio and can be found open-source at the **link**.
 
@@
 
+Eighteen outcrops (primary and secondary) were revisited or identified in the Algarve region, nine in the westernmost territory, and nine to the east (between Loulé and Tavira).
+From these, 70 samples were recovered and analyzed.
+Some of these samples were recovered as isolated finds or in secondary settings.
+
+On the westernmost part of the Algarve, there are mainly cherts from two different formations: Lower Jurassic and Upper Jurassic.
+The latter can be found in a single known outcrop - Praia da Mareta.
+Lower Jurassic outcrops are more common and have been better studied (ribeiro_evolucao_2005).
+These outcrops are heterogeneous, showing different characteristics and chert colors.
+
+The Lower Jurassic cherts can be grouped in three main macroscopic types: yellow and/or red type; grey/brown type; yellow and red with fossils type.
+These macroscopic types not only show color differences but also in the fossil content, visible to the naked eye and stereomicroscope.
+The first two types are present in all outcrops.
+They are mainly characterized by dull to medium luster and opaque translucency, although some samples were sub-translucent.
+The feel ranges between smooth and semi-smooth, although many of the Belixe cherts have a rough feel.
+In the Yellow/Red cherts, fossil content is present but visible only as white, red, or yellow speckling.
+The Grey/Brown cherts show little fossil content, barely visible with the stereomicroscope.
+The Yellow/Red with fossils cherts show a large number of big fossils, which are easily seen by the naked eye and can be identified under the stereomicroscope.
+According to the microscopic approach, the Lower Jurassic cherts of Western Algarve are composed mainly of microcrystalline quartz, with textures that range mostly from wackestone to packstone.
+In more than 50% of the samples, no fossil can be identified, as all fossils, albeit common to very frequent, are poorly preserved, filled with chalcedony or quartz, and without any identifiable morphology.
+Whenever identifiable, fossils present in the sample are Echinoderms, Radiolarians, Sponge spicules, and a Bivalve shell.
+
+Despite the similar characteristics between these cherts, independent of color or geographical location, the outcrops are heterogeneous and show varying characteristics, which may be of importance to distinguish between chert sources within the Lower Jurassic formation.
+These outcrops have been divided into four groups, following the available literature: Cabo de S.
+Vicente (including Cabo de S.
+Vicente and Aspa); Foz dos Fornos; Ponta dos Altos; Praia do Belixe (which includes Belixe Sul); and Ferrel.
+
+The Cabo de S.
+Vicente (CSV) and Aspa (ASP) chert is characterized by abundant nodules in the natural rock banks of the cliffs, appearing as horizontal layers within the parent rock.
+The banks seem to be mainly dolomite or dolomitic limestones.
+The process of dolomitization seems to have affected the chert nodules, as they often present different levels of silicification from the peripheral areas of the nodule to the interior (which is more silicified), which also affects the knapping quality.
+The nodules vary in size, ranging from small 4 cm in diameter circular nodules to bed-like groups of nodules of ~20 cm in width.
+At the Aspa outcrops, the nodules are less frequent and smaller.
+Due to the proximity to the cliffs, the visibility of the chert nodules is good, and in present times, small chunks of chert (without cortex or with small amounts of parent rock attached) accumulate in secondary deposition nearby.
+Foz dos Fornos (FZF) and Ponta dos Altos (PdA) show similarities to the CSV outcrops.
+The nodules are visible in several banks of dolomite, dolomitic limestone, and limestone, partially covered by soil.
+The nodules can be circular, around 5 cm in diameter, or wide nearly 20 cm in width.
+Despite their size, these cherts are frequently filled with fractures that fragment the larger nodules into smaller volumes of raw material.
+Alike CSV, FZF, and PdA also show cherts with differing degrees of dolomitization, although in apparent smaller quantities than CSV.
+Besides the abundant presence of primary outcrops, there are also abundant chert nodule fragments in secondary deposition, down the slope of the cliff (in the case of FzF) or at the top of the cliff, on a sand path (in the case of PdA).
+These are small, between 1-4 cm in width, but of easy access.
+Between the FZF chert and the PdA, the main differences seem to be the cortex and parent rock, which show differing reactions to hydrochloric acid, the first being dolomite or dolomitic limestone, and the second being mostly limestone, with some degree of dolomitization in certain areas.
+
+Praia do Belixe (PBLX) is characterized by the abundance of chert nodules throughout the dolomite layers of the cliff area.
+They are visible in certain areas of the cliff and within the rock shelters.
+The nodules can be small, around 5 cm in diameter, sometimes reaching more than ~30 cm in width, or bedded, as chert layers between the dolomite layers.
+The cherts show varying degrees of dolomitization and are mostly characterized by coarse to semi-smooth feel, dull luster, and a medium to low knapping quality.
+Unlike the other outcrops, no chert nodule fragments were found close to the cliffs, and samples could only be recovered directly from the embedded nodules in the cliff walls.
+Nodules scattered on the floor were only located at Belixe Sul (BLS), a primary outcrop nearly destroyed located on a field, north of the beach area.
+The chert in this outcrop showed no differences from PBLX, aside from the size of the nodules, which were smaller and often showed signs of post-depositional alterations.
+
+The third location for chert has been previously identified north of BLS.
+Belixe Norte (BLN) is located on a dirt road and unused agriculture field.
+Several chert fragments were collected in this location.
+However, BLN is in proximity to an archaeological site and several collected samples were lithic artifacts.
+No larger nodules or outcrop were identified in this location.
+The samples recovered from the location also seem to corroborate that BLN should not be considered an outcrop, as they do not match the local cherts and rather, resemble most of the samples recovered from the Eastern Algarve.
+Ferrel, unlike the other outcrops, is located inland and away from the coast.
+Due to its location in a homonymous village, the state of the outcrop is poor, and all samples were either recovered as scattered nodules or from larger blocks of rock, from a partially destroyed outcrop.
+The proximity of an archaeological site nearby also raises questions regarding the nodules found in secondary deposition, as these may be surface finds.
+Despite these caveats, the recovered samples are similar to those from the other outcrops, albeit with better quality, being characterized by a shiny to medium luster and smooth to semi-smooth feel.
+All surface fragments and nodules were small, with around 2 to 3 cm of width which may be explained by the state of the outcrop.
+Unlike the outcrops from the Lower Jurassic of the western Algarve, this region of the Algarve has only one identified outcrop for Upper Jurassic cherts.
+These are located at Praia da Mareta (MAR) and abundant, or in a secondary deposition at Ponta da Atalaia (PtA).
+At Praia da Mareta the nodules are only easily accessible on the beach, where large chunks of the cliff (~1 m in diameter) are transported by the waves.
+Several chert nodules of different sizes can be found in the parent rock washed ashore, ranging between 2 cm to 20 cm in diameter.
+The quality of the chert also varies, possibly related to different dolomitization stages of the nodules, although this may also be influenced by chemical and physical alterations to the chert.
+At Ponta da Atalaia the chert can be found atop the cliffs, with rare nodules scattered on the floor.
+The Upper Jurassic cherts are very similar to the Lower Jurassic, with dull to medium luster and grey/purple colors.
+The translucency ranges from opaque to areas where the chert is translucent.
+This translucency may be a significant difference to distinguish between outcrops.
+Petrographically, the cherts are also similar to the Lower Jurassic ones.
+The only identifiable difference is the presence of calcispheres.
+All samples from the Mareta outcrops seen under the petrographic microscope showed the presence of abundant calcispheres , which is not always apparent with the stereomicroscope.
+Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic, which was uncommon and scattered at the top of the cliffs by the beach.
+
+On the eastern part of the Algarve, chert-bearing known formations are from the Middle to Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively.
+
+The Malhão formation chert (dating to the Middle Jurassic) was identified in three outcrops, in the Faro parish and the Tavira parish.
+Whenever in a primary outcrop, this chert was homogeneous.
+The secondary deposits were recent waterlines and slope deposits, and the cherts were often characterized by intense post-depositional alterations.
+In these cases, it was not possible to confirm the outcrop location.
+In these outcrops, the nodule frequency varied from common to abundant.
+The nodules are roundish, ranging between 3 to 5 cm of maximum width.
+In all cases, access to the outcrops was easy.
+Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits.
+The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts.
+In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel.
+They are easily identifiable through the high amounts of macroscopically visible inclusions, which look like white speckling in plain sight.
+Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified.
+The petrographic analysis shows for the Malhão cherts from Casal da Colina (primary and secondary outcrops) have high amounts of dolomite within the chert.
+All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all poorly preserved and replaced by chalcedony or quartz).
+These fossils are Sponge spicules, Radiolarians, Ostracods, Tentaculites (also previously identified by – ), Echinoderms, and Calcispheres.
+
+The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve, above Moncarapacho in the Olhão parish.
+Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock.
+No chert was identified in any secondary deposits, which might be related to the anthropic alteration of the landscape.
+Smaller nodules broken from the parent rock were identified near the primary source in a field.
+Whenever embedded in the parent rock, the nodules varied in size (~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock.
+The cherts show little macroscopic variability between nodule and outcrop.
+They are grey/brown (with visible yellow inclusions).
+Within nodule, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas.
+Some of the nodules also show the variability of translucency, with translucent areas, with a very fine grain, and little presence of visible inclusions.
+The petrographic analysis shows that the cherts range from a wackestone to packstone texture, which was already seen macroscopically.
+They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals.
+Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shells, Sponge spicules, Ostracod, Echinoderms, and Gastropod.
+Perhaps add a paragraph about the sources which were marked on the geological map but could not be located.
+
 # Discussion
 
+The prospection works and analyses of the collected geological samples show that the south of Portugal has a high potential for raw material studies, especially regarding chert.
+The presence of chert-bearing outcrops in the westernmost part of the Algarve, center, and east would provide several possibilities for sourcing and procurement whenever groups moved throughout the territory.
+This is further important when we consider the geology of this territory.
+The geology of the Algarve itself may have played an important part in how groups procured their raw materials, specifically, their chert, a task that has been identified as essential for hunter-gatherer groups.
+To the south, communities would only have access to chert-bearing outcrops up to the coast.
+To the north, the mountain range would not only have provided no chert nodules but may have also hampered the movement of populations, forcing groups to move east and west instead of north or south.
+This movement may have facilitated the gathering of cherts from different formations within the Algarve, posteriorly then brought into the sites.
+Especially for Middle and Upper Paleolithic occupations, understanding the sources of chert in the Algarve may provide data about where in the territory they were sourcing their raw materials, and how they were using the territory when their movement was constrained by the Peninsula’s natural barriers.
+This question remains unanswered, but raw material data and comparative studies between archaeological assemblages and geological samples may provide the necessary results to answer it.
+
+However, tracking these movements and procurement patterns is only possible if the cherts from the different formations and outcrops can be traced back to their sources.
+This presented itself as the first caveat for this type of study, since for the Algarve, for example, all cherts come from Jurassic formations in pelagic environments.
+Despite the similar formation environments, in general, there seem to be relevant differences between the cherts of different formations and periods.
+This is further relevant given the fact that they are geographically distant.
+Within formations, however, there are no discernable differences, both at a macroscopic and petrographic level, as these do not seem to be useful to distinguish between outcrops.
+That is most obvious in the Lower Jurassic outcrops of west Algarve.
+The identified chert groups, which varied mostly in color and fossil content, are present in several outcrops.
+In this region, the variables which may be better used to understand which outcrops were visited may be the quality and size of the nodules.
+The latter, for example, is an important variable in the Belixe outcrops, which show the largest volumes of rock, even if the knapping quality is worse than some other available, smaller nodules.
+Size may be used in conjunction with other technological data, to understand whether different nodules were being explored differently based on their size, or their procurement was being preferred over other smaller nodules in possibly closer outcrops in the region.
+Size may also be an important variable to distinguish between nodules collected in primary or secondary outcrops.
+Whenever found in secondary outcrops (without the presence of parent rock and with rounded and/or altered cortex) the nodules were small and altered.
+Although in other regions, secondary outcrops such as river beds, for example, may be of extreme importance to raw material procurement, it seems that for the regional cherts, these would be procured in primary outcrops.
+
+The Upper Jurassic nodules of the western Algarve also show larger volumes than those from the Lower Jurassic.
+This, together with the translucency, which is often characterized by sub-translucent fabrics, may be a good macroscopic indicator of the chert’s provenance.
+However, for a reliable distinction between the Lower Jurassic and the Upper Jurassic cherts, petrographic analyses and the identification of calcispheres may be necessary.
+The differences identified between the cherts of the various formations can be seen both at a macroscopic and petrographic level.
+Given the formation setting, petrographically, all the cherts from the Algarve are fairly homogeneous - marine origin, in limestone or dolomitic limestone rocks, all formed during the Jurassic.
+The use of fossils for the identification of the cherts is also difficult since these are often not well preserved enough to allow the identification of species that may connect a group of cherts.
+The size, frequency, and preservation state of the fossils seem to be, then, one of the defining criteria for discerning cherts from different formations, and thus, different geographic areas.
+These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana, and Malhão - to be differentiated without the need for thin sections.
+This is especially important for archaeological collections, especially those which may be small, with small archaeological pieces, or for the study of older collections to which other means of analysis may not be available.
+
+This data seems to confirm the potential of a macroscopic analysis to study the cherts of the Algarve.
+Albeit applying different methodologies, such as petrographic analyses, to these cherts is a way of completing the petrographic study of a collection, reliably applying mostly a macroscopic analysis to the assemblages coming from southern Portugal helps tackle issues such as the destructiveness, costliness, and time consumption of some methods.
+The research was able to complete the collection, providing samples of the available cherts which may be compared with archaeological samples.
+However, there are some caveats to this type of comparative studies.
+Landscapes have changed through time, especially with the influence of modern societies.
+House constructions, agricultural fields, and roads, for example, have modified the landscape, possibly altering the availability and visibility of raw materials.
+The existence of chert at specific points on the map may not necessarily mean that they do not exist elsewhere.
+However, identifying characteristics that are common to most, if not all cherts of a region, may be a good indication that some cherts are not local.
+
+Another caveat regarding chert sources, especially in a geographic area like the Algarve, is the possibility of some outcrops being submerged.
+Previous studies have identified the existence of Jurassic outcrops with chert nodules on the west coast, submerged by water.
+In times when the sea level was similar to the current one, these outcrops would not have been accessible, even during low tide, the only possibility for their procurement in nodules dragged to the shore.
+However, during periods where the sea level was lower due to water freezing in the polar caps (during the LGM, for example), large portions of the coast would have been accessible.
+This raises the question of whether the chert reference collection is representative of all the variability of chert in the region, including the submerged coastal areas.
+For studies focusing on periods like the Proto-Solutrean and Solutrean in the Algarve, for example, where nearly 20 km of the shore was exposed at the peak of the LGM, there may have been the exploitation of unknown sources or cherts with more variability than that known from the recovered samples.
+Despite the caveat, this raises the possibility to understand whether this new portion of landmass altered the raw material procurement patterns of these groups, or added new resources which had been previously unavailable.
+Studies that compare Gravettian and Magdalenian assemblages to Proto-Solutrean and Solutrean assemblages, within one single site, may give new insights to this question.
+
 # Conclusion
+
+The research and results in this paper continued and completed an important study of chert raw materials in southern Portugal.
+The results show the presence of four different chert formations, dispersed in the western and eastern Algarve.
+Within formations, there is a lot of variability in the nodules and the outcrops.
+There are however identifiable macroscopic and petrographic differences between formations.
+** Although the petrographic analysis is essential to identify the fossils present in the chert, a macroscopic approach seems to be pertinent for a quick and inexpensive analysis to distinguish between cherts of the different formations.
 

### 2022-10-04 analysis/supplementary-materials/SP6.Rmd
- Uploading the word draft with corrections

---
title: "Untitled"
author: "Joana Belmiro"
date: '2022-10-01'
output: pdf_document
---

# Sample ID and provenance

Sample ID: SP6_CSV

Lithology: Chert

Collection: LusoLit

Outcrop: Cabo S. Vicente

Unit/facies: Lower Jurassic

***

## Macroscopic description

### **Color**

The color distribution is Single and the color is Light gray (2.5Y 7/2).

### **Fabric**

The luster is Dull and the translucency is Opaque. The feel is Semi-smooth to Rough and the grain is fine. The distribution of the fabric is Even.

### **Inclusions and fossil content**

### **Cortex**

The cortex is Outcrop type with a Gradual transition.

### **Quality**

The fracture is Conchoidal with Fractures. The knapping quality is Medium.

### **Observations**

***

## Petrography analysis form

### Textural classification

**Texture**
Mudstone

**Microstructure**
Massive

***

### Composition

| Orthochem       | Type | %   | Description                             |
| --------------- | ---- | --- | --------------------------------------- |
| MiC quartz (gr) | SE   | 90  |                                         |
| Dolomite        | SE   | 10  | Dolomite crystals have different sizes. |

| Allochem | Freq     | Description |
| -------- | -------- | ----------- |
| Oxides   | Uncommon |             |

| Bioclasts  | Freq     | Description |
| ---------- | -------- | ----------- |
| Ghosts     | Uncommon |             |
| Echinoderm | Uncommon |             |

### Other textural characteristics

**Porosity**

Total porosity (%): 5

Porosity: Vuggy

***

### Photos

| Image ID | Aug. | Description                                                           |
| -------- | ---- | --------------------------------------------------------------------- |
| SP6_001  | 10x  | General view of oxides and a possible Echinoderm spine.               |
| SP6_002  | 10x  | Ghosts with quartz filling.                                           |
| SP6_003  | 10x  | Limit between chert and dolomitic limestone (cortex.)                 |
| SP6_004  | 10x  | Possible Echinoderm spine.                                            |
| SP6_005  | 20x  | Detail of rhomboid dolomite crystals close to the edge of the sample. |

***
### Observations

- Ghosts are filled with microcrystalline quartz and poorly preserved.
- Dolomite is visible in the chert area closer to the cortex area.

***
### Analysis information

- Analyst: JB

- Date: 02.23.2022

- Equipment: Leica DM2500 P

### 2022-10-06 analysis/supplementary-materials/petrography_dataset.Rmd
- Petrography dataset RMD and PDF

---
title: "Petrography dataset"
output: pdf_document
---

```{r setup, include=FALSE}

library(knitr)
library(kableExtra)

```

```{r petrodataset, echo=FALSE}

petro_dataset <- data.frame(Variable = c("ID", "Outcrop/level", "Lithology", "Texture", "Microstructure", "Orthochem", "Orthochem type", "Orthochem description", "Orthochem (%)", "Allochem", "Allochem (freq)", "Bioclast", "Bioclast (freq)", "Porosity (%)", "Porosity type", "Sedimentary structures"),
                            "Allowed Values" = c("","","", "Mudstone, Wackestone, Packstone, Grainstone, Boundstone, Other", "Homogeneous, Banded, Laminar, Nodular, Brexoid, Other", "", "Essential (ES), Accessory (AC), Secondary (SE)", "", "", "", "Rare, Uncommon, Common, Very frequent", "", "Rare, Uncommon, Common, Very frequent", "", "Interparticle, Moldic, Fenestral, Fracture, Vuggy, Shelter, Other", "Parallel lamination, Convoluted lamination, Bands/zonations, Burrow, Other"),
                            Description = c("Individual thinsection ID.", 
                            "Outcrop name or archaeological level.", 
                            "Type of rock", 
                            "Mudstone: Muddy carbonate rock containing less then 10 % grains; Wackestone: Mud-supported carbonate rock containing more than 10 % grains; Packstone: Grain-supported muddy carbonate rock; Grainstone: Mud-free carbonate rocks, which are grain supported; Boundstone: Carbonate rocks showing signs of being bound during deposition.", 
                            "Distribution of crystals and clasts within the rock at a microscopic scale. Homogeneous: equally spread in the rock; Banded: distributed in bands; Nodular: distributed in clumps; Brechoid: fracturing of the rock irregularly.", 
                            "Materials formed in two ways: 1) deposited directly from supersaturated aqueous solutions due to chemical reactions or evaporation; 2) formed by the replacement of existing sedimentary materials.", 
                            "Essential: minerals that form more than 5% of the volume of the rock; Accessory: Minerals with proportion of less than 5% of the volume of the rock; Secondary: products of the alteration (hydrothermal or physical), independent of the proportion within the rock.", 
                            "General description of the orthochem and where it is identified.", 
                            "Approximate percentage of the orthochem's presence in the total thin-section area.", 
                            "Material formed by the movement and reorganization into new shapes by chemical, physical or biological processes within the depositional basin (ex. ooliths, fecal pellets, iron oxide minerals).", 
                            "Rare: present one or two elements; Uncommon: present three to 10 elements; Common: present 11 to 20 elements; Very frequent:> 20 elements.", 
                            "Also known as skeletal particles, are the remains (complete or fragmented) of the hard parts of carbonate-secreeting organisms.", 
                            "Rare: present one or two elements; Uncommon: present three to 10 elements; Common: present 11 to 20 elements; Very frequent:> 20 elements.", 
                            "Approximate frequency of effective porosity.", 
                            "Interparticle: porosity between particles; Moldic: Porosity formed by selective removal of an individual constituent of the rock; Fenestral: Pores larger than grain-supported interstices (interparticle); Fracture: Porosity formed by fracturing; Vug: Pores larger then 1/16 mm in diameter and somewhat equant in shape; Shelter: Porosity created by the sheltering effect of large sedimentary particles; Other: Other types of porosity.", 
                            "Parallel lamination: sedimentary strata less than 10 mm thick, recognizable due to variation in structure or composition and more or less parallel bounding surfaces; Convoluted lamination: symmetrical about a vertical plane or leaning and asymmetrical, and usually exhibit narrow vertical upturned laminae, often truncated at the top, separated by a broader synclinal downfolds; Bands/zonations: limited areas with different characteristics related to changes in the sedimentation or cementation process; Burrow: bioturbation structures caused by activity of an organism that disrupts the stratification features; Other: Other types of sedimentary structures."),
                            Reference = c("", "", "", 
                            "According to Dunham (1962)", 
                            "According to Dorado (1989, pp. 21)", 
                            "According to Vernon (2018, pp. 24-25)", 
                            "According to Dorado (1989, pp. 26)", 
                            "", 
                            "", 
                            "According to Vernon (2018, pp. 25, 27)", 
                            "", 
                            "According to Adams, McKenzie and Guilford (1991, pp. 39)", 
                            "", 
                            "", 
                            "According to Choquette and Pray (1970)", 
                            "According to Middleton et al. (2003)")
                            )
knitr::kable(head(petro_dataset[1:4], 16), booktabs=TRUE, caption = "Dataset for petrographic analysis") %>% 
  column_spec(1, bold = T, width = "3 cm") %>% 
  kable_styling(latex_options = "scale_down") %>% 
  kableExtra::landscape() %>% 
  column_spec(2, width = "4 cm") %>% 
  column_spec(3, width = "15 cm") %>% 
  column_spec(4, width = "4 cm")

```

### 2022-10-08 analysis/supplementary-materials/thinsection_list.Rmd
Supplementary data: - geology references - macroscopic analysis table - thin section list

---
title: "List of thin sections"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

| Sample ID | Type               | Laboratory | Outcrop          | Sample collection |
| --------- | ------------------ | ---------- | ---------------- | ----------------- |
| SP6       | Covered            | UB         | Cabo S. Vicente  | 2021              |
| SP7       | Covered            | UB         | Cabo S. Vicente  | 2021              |
| SP9       | Covered            | UB         | Ponta dos Altos  | 2021              |
| SP10      | Covered            | UB         | Ponta dos Altos  | 2021              |
| SP14      | Covered            | UB         | Praia Belixe     | 2021              |
| SP15      | Covered            | UB         | Praia Belixe     | 2021              |
| SP18      | Covered            | UB         | Praia Belixe     | 2021              |
| SP21      | Covered            | UB         | Belixe Sul       | 2021              |
| SP23      | Covered            | UB         | Belixe Norte     | 2021              |
| SP24      | Covered            | UB         | Cabo S. Vicente  | 2021              |
| SP27      | Covered            | UB         | Cabo S. Vicente  | 2021              |
| SP28      | Covered            | UB         | Aspa             | 2021              |
| SP32      | Covered            | UB         | Foz dos Fornos   | 2021              |
| SP33      | Covered            | UB         | Foz dos Fornos   | 2021              |
| SP34      | Covered            | UB         | Ponta dos Altos  | 2021              |
| SP34      | Covered            | UB         | Ponta dos Altos  | 2021              |
| SP36      | Covered            | UB         | Ponta da Atalaia | 2021              |
| SP39      | Covered            | UB         | Andorinha        | 2021              |
| SP40      | Covered            | UB         | Ferrel           | 2021              |
| SP42      | Covered            | UB         | Boca do Rio      | 2021              |
| SP59      | Covered            | TSL        | Jordana          | 2022              |
| SP61      | Covered            | TSL        | Peral            | 2022              |
| SP56      | Covered            | TSL        | Jordana          | 2022              |
| SP54      | Covered            | TSL        | Guilhim          | 2022              |
| SP53      | Covered            | TSL        | Guilhim          | 2022              |
| SP58      | Covered            | TSL        | Jordana          | 2022              |
| SP55      | Covered            | TSL        | Caliços          | 2022              |
| SP50      | Covered            | TSL        | Casal da Colina  | 2022              |
| SP52      | Covered            | TSL        | Casal da Colina  | 2022              |
| RT82      | Covered            | TSL        | Praia Belixe     | <2021             |

### 2022-10-24 analysis/paper/paper.Rmd
- Corrected draft - Added photos used for plates

@@
 library(sf)
+library(raster)
+library(spData)
 library(rnaturalearth)
@@
 library(tidyr)
+library(tmap)
+
+install.packages("tmap")
 
@@
   rename(SampleID = NewCode)
+
+```
+
+```{r map, echo=FALSE}
 
@@
   geom_sf() +
-  geom_point(data = chertfield, aes(x = long, y = lat, shape = State, color = State), size = 4) +
+  geom_point(data = chertfield, aes(x = long, y = lat, shape = State, color = Age), size = 4) +
   #geom_label_repel(data = chertsamples, aes(long, lat, label=NewCode), box.padding = 0.5, max.overlaps = 30) +

### 2022-12-21 analysis/paper/paper.Rmd
- Commit after hard reset

--- 2022-09-20 analysis/paper/paper.Rmd
@@
   - name: Joana Belmiro
-    email: fl@oneplace.edu
-    institute: [UofO]
+    email: jfbelmiro@ualg.pt
+    institute: ICArEHB
     correspondence: true
   - name: João Cascalheira
-    email: fl@another.edu
-    institute: [UofA]
+    email: 
+    institute: ICArEHB
+    correspondence: false
+  - name: Xavier Terradas
+    email: 
+    institute: CSIC
     correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - ICArEHB: Universidade do Algarve
+  - CSIC: Bleh
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-library(ggplot2)
-theme_set(theme_bw())
-library(sf)
-library(rnaturalearth)
-library(rnaturalearthdata)
 library(readr)
-library(ggspatial)
-library(ggrepel)
 library(dplyr)
 library(kableExtra)
-library(flextable)
-library(data.table)
-library(tidyr)
-
-```
-
-```{r databases, include=FALSE}
-
-petrography <- read_csv(here::here('analysis/data/raw_data/PetrographyDB.csv')) # Reading the petrography data
-chertfield <- read_csv(here::here('analysis/data/raw_data/chert_field.csv')) # Reading the field data
-
-fielddata <- chertfield %>% # Cleaning the Field data
-  mutate(Geography = case_when(long <= -8.2 ~ "Ocidental",
-                               long > -8.2 ~ "Oriental")) %>% 
-  filter(`Raw Material` == "Chert") %>% 
-  rename(SampleID = NewCode)
-
-```
-
-```{r map setup, include=FALSE}
-
-sf::sf_use_s2(FALSE)
-world <- ne_countries(scale = "medium", returnclass = "sf")
-#europe <- world[world$continent == "Europe",]
-europe_cropped <- st_crop(world, xmin = -15, xmax = -7.5,
-                                    ymin = 35, ymax = 37.6)
-
-```
-
-```{r map, echo=FALSE, include=FALSE}
-
-ggplot(data = europe_cropped) +
-  geom_sf() +
-  geom_point(data = chertfield, aes(x = long, y = lat, shape = State, color = State), size = 4) +
-  #geom_label_repel(data = chertsamples, aes(long, lat, label=NewCode), box.padding = 0.5, max.overlaps = 30) +
-  annotation_scale(location = "br", width_hint = 0.5) +
-  annotation_north_arrow(location = "br", which_north = "true", 
-        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
-        style = north_arrow_fancy_orienteering)
-  #coord_sf(xlim = c(long_min, long_max), ylim = c(lat_min,lat_max), expand = FALSE)
-
-```
-
-```{r databasesetup, include=FALSE}
-
-CStable <- chertfield %>% 
-  select(NewCode, Site, `Sample variability`, State, Deposit, Type, Frequence, `State of site`, `Thin-sections`)
-
-```
-
-```{r cstable, echo=FALSE, include=FALSE}
-
-regulartable(CStable) %>% 
-  theme_zebra()
+library(knitr)
 
@@
 
-The South Iberian territory is a key area to understand cultural transitions from the Middle Paleolithic to early Prehistory .
-The Algarve region specifically (south Portugal) has contributed greatly to understanding pre-history at a European level, with several well-known archaeological sites and publications.
-As a territory located at the tip of the European continent and with a generally warm climate even during the coldest periods which characterize the Late Pleistocene, it has been regarded as one of the most significant glacial refugia in Europe (Hewitt, 2000; Gómez and Lunt, 2007; González-Sampériz et al.
-2010).
-Several authors have suggested that the Middle to Upper Paleolithic cultural sequences present in the Iberian Peninsula territory are related to adaptation and reorganization processes related to abrupt climatic events (Brädtmoller et al., 2012; Cascalheira et al., 2021).
-As such, the south Iberian territory witnessed several moments of crisis, cultural changes, and behavioral adaptations.
-An example of this is the late occupation of the territory by Neanderthal populations and their consequent disappearance.
-The Iberian record for the Middle Paleolithic allows for nearly 250 kyr of Neanderthal occupations to be studied until the extinction interval (de La Torre, Martínez-Moreno, and Mora, 2013).
-Several studies suggest that Neanderthal presence in southern Iberia lasted until c.
-37 ka cal BP or later, with a different trajectory than that found in other places of Europe (Cascalheira, Gonçalves, and Maio, 2021).
-
-The several cultural shifts of the Upper Paleolithic (UP) are also represented in the region, in sites with long chronological occupations, from the first UP occupations such as the Aurignacian (not detected in southern Portugal) and the Gravettian, such as the archaeological site of Vale Boi .
-This site, containing one of the most complete UP sequences of the Iberian Peninsula, offers the opportunity to understand the movements of the first groups of Humans to reach the territory, their cultural changes throughout the Upper Paleolithic, as well as the impact of abrupt climatic change in culture, such as the Heinrich Event 2 (HE 2) and the Last Glacial Maximum (LGM).
-Finally, south Iberia also shows the existence of several Mesolithic, Neolithic, and Chalcolithic settlements and occupations.
-In south Portugal, for example, studies have shown the use of riverine and estuarine areas for Mesolithic settlements (Carvalho, 2008) and a concentration of sites by the westernmost coast of the Algarve (Bicho, 2009), as well as clear settlement and cultural shifts in the Neolithic occupations, followed by Chalcolithic occupations (Sánchez et al., 2012).
-This complete recent prehistory chronology makes southern Iberia, and in specific, southern Portugal, an ideal territory to study the replacement of the last hunter-gatherers of Europe by the first farmers, and the subsequent transformations these communities went through.
-
-From at least ~40 000 cal BP to 3 000 cal BP (** , Valera, 2012), South Portugal has seen moments of crisis and cultural change of different kinds, which gives this territory a large potential for prehistorical archaeology studies and the study of human behavior in a wide chronological range.
-With such a wide chronology, differences in the exploitation of resources, management of territories, settlement patterns, subsistence techniques, climate, technology, and culture, it is important, however, to find a unifying subject that can be studied, analyzed, and compared, bringing together several studies and sites to create a more unified story of the evolution of human behavior.
-That subject may be the raw material that was used by Neanderthals, Human hunter-gatherers, and farmers - that raw material is chert.
-Whenever geologically present (and at times, even when it is not geologically present), Chert has served as a raw material for the production of blanks to more specialized and complex tools.
-As a good quality rock, it was procured from close to far-away locations, transported, knapped, retouched, re-used, heated, and sometimes traded, traveling long distances in the hands of different groups which produced them.
-Countless archaeologists and studies have shown the potential of this raw material to understand cultural aspects such as foraging strategies and territory use, mobility, shifts in technological necessities, cultural preferences, changes in raw material visibility and accessibility, or even the establishment, expansion, and shrinkage of social networks ** .
-As a raw material, chert has been studied in South Portugal.
-The study of chert-bearing outcrops has been a topic of interest for both geologists and archaeologists in the Algarve region, albeit for different reasons.
-Geological literature has often focused on the processes that formed the seas and landmasses of the Algarve basins, as well as the paleobiology and paleobiogeography of the region.
-Archaeological literature has focused on chert sources as a way to predict the availability of mineral resources, and the existence of settlements, characterize hunter-gatherer culture throughout prehistory and understand group mobility and social exchange through time.
-Geological studies in the Algarve region go back to the 19th century and throughout all of the 20th century, through the creation of geological cartography.
-These first works provided the basis for the geological knowledge of the Algarve basin, and it allowed further studies within the geosciences, as a large number of Doctoral theses.
-
-Unlike what happens in other areas of Portugal (add references from Aubry, Abrunhosa, etc), archaeological studies focusing on raw materials, especially chert, have been more scarce.
-In general, lithic technology studies in the Algarve have approached raw materials through a macroscopic methodology, often related to the location of the settlement (i.e., [@bicho_o_2003]; Pereira et al 2016 ) or the description of the collection (i.e., [@cascalheira_tecnologia_2010], [@marreiros_lithic_2015]).
-The work of Veríssimo (2005) focusing on the occurrence of chert in western Algarve provided the initial basis for comparative studies, maintaining however solely a macroscopic component.
-Finally, the creation of LusoLit (Telmo et al.
-2016) and the collection of samples from the Algarve provided a new leap in the study of chert in the region.
-Given the existence of several chert-bearing outcrops in South Portugal, preliminary raw material studies, a region of archaeological importance, and the existence of a rich archaeological record, the Algarve stands as a region full of potential for studies of chert and the development of a lithotheque which may serve for regional studies, but also for comparative studies outside, where chert may have been transported and traded, offering the chance to complete the record of exotic cherts which may be coming from the Algarve.
-
-This paper aims to locate and characterize the nature of chert sources in southern Portugal and their availability.
-Furthermore, it aims to test the potential of different methodological approaches to characterize the regional cherts and develop a methodological approach adapted to this study area.
+Southern and Western Iberia have often been considered key areas to understanding techno-cultural transitions from the Middle Paleolithic to the end of the Upper Paleolithic.
+As territories located at the western tip of the European continent and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, these have been regarded as some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022].
+For this reason, Southernwestern Iberia has frequently been at the center of some of the most debated topics regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017].
+A particularly good example is the region's role in possibly being one of the last territories to be occupied by Neanderthal populations right before their complete disappearance [@finlayson2006; @zilhão2017].
+Neanderthals occupied the European continent for more than 300.000 years and are thought to have disappeared while modern humans arrived on the territory [@trinkaus1979; @mellars2004; @finlayson2006; @jennings2011; @rogers2017].
+The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c.
+37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022] .
+This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018].
+Another example is Southwestern Iberia's importance in the discussion of Upper Paleolithic technocomplexes transitions.
+Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011].
+The Heinrich Event 2 (HE 2) at the onset of the Last Glacial Maximum (LGM), for example, is a period marked by important social and technological transformations.
+This climatic event was characterized by abrupt and drastic climatic changes that impacted human behavior all across westernmost Europe [@gamble2004; @sanchezgoñi2010].
+The identification of a Proto-Solutrean phase in central and southern Portugal with a very distinct index fossil (the Vale Comprido point), and its direct association with the HE 2, put these regions amongst some of the most important case studies of how environmental dynamics have affected human adaptations during the last glacial.
+Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017].
+Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources.
+This continuous use would then create long-term regional adaptative structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches.
+In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012].
+These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints.
+One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi.
+This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of Iberia [@bicho2013; @cascalheira2013; @cascalheira2017].
+Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017].
+Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials.
+The maintenancee of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016].
+The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012].
+A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on macroscopic preliminary results or the differentiation between possibly local and non-local sorts of raw materials.
+Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003].
+As an essential part of late Pleistocene hunter-gatherer adaptations, lithic raw materials have the potential to provide insights into the adaptive strategies of those populations [@féblot-augustins1993; @kuhn1995; @mellars1996], mostly regarding land-use, technological organization, but also cultural and social interactions.
+In fact, the selection and procurement of raw material has been suggested as a key stage for the technological organization of hunter gatherer groups [@bamforth1997].
+Thus, changes in the frequencies of raw materials within the archaeological records may provide evidence for changes in that organization, or even resilience of specific choices, all of which can reflect culturally transmitted preferences within a group of hunter gatherers [@brown1999].
+Several models, both formal and informal, have shown that changes of raw material in the archaeological record may reflect different procurement strategies [@binford1985; @binford1979; @gould1985; @gouldsaggers_1985], either related to mobility strategies [@ambrose1990; @kuhn2004; @kuhn1991; @mccall2007] or changes in the availability of raw materials, possibly related to environmental change [@brantingham2003; @brown1999].
+Raw materials are also closely linked to the tools produced with them -- the costs of manufacturing technology can be related to the availability of raw materials in the landscape [@torrence1983; @bousman1993; @parry1987; @mackay2011], or the characteristics of the raw materials which might relate to functionality [@mackay2008; @minichillo2006; @stout2002], through the choice of raw materials for specific technologies [@brown1999].
+Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999].
+In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like Vale Boi, it is necessary to know the landscape and the raw material sources available in the territory.
+To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage.
+This is true for most of the informal models which try to understand why raw materials change in the archaeological record.
+Furthermore, formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015].
+In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022].
+In southern Portugal previous efforts have been made to create such a database.
+The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from Vale Boi based on macroscopic approaches.
+The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region.
+A few geological studies have also contributed to understand the availability and characteristics of chert in southern Portugal [i.e., @rocha1976; @ribeiro2005].
+Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages.
+Furthermore, a large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists.
+As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation.
+Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity and availability of chert raw materials in the southernmost region of Portugal -- the Algarve.
+Our main goal is to establish a reference for chert raw materials in an understudied region in what regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere.
+This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts.
+The data presented here is integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.
 
@@
 
-The Algarve region is located in southern Portugal, framed north and east by the Alentejo region and Spain respectively.
+The Algarve is the southernmost region of Portugal, framed north by the Alentejo region and east by Spain.
 To the west and south, it is bordered by the Atlantic ocean.
-Geologically, it is composed of two main geological units: the South Portuguese Zone (SPZ) and the Algarve basin.
-The SPZ is composed of four “domains” of different lithologies and ages, extending up to Alentejo, although all within the Paleozoic period.
-It is overlain unconformably by the Mesozoic sedimentary rocks of the basin [@fernandes_new_2012].
-The Algarve basin is one of the broader Jurassic outcrops in Portugal (Malchus and Kuss, 1988).
-It corresponds to the Mesozoic-cenozoic sediments that outcrop south of Portugal, from Cape of St.
-Vicent to Vila Real de St.
-António.
-The Algarve basin is associated with the opening of the central Atlantic and with the eventual oceanic crust formation in the western part of the Tethys sea, between the Algarve and North Africa (Terrinha et al., 2013).
-After the sedimentation period between the Triassic and the Sinemurian (Lower Jurassic), the Mesozoic basin was divided into two sub-basins (western sub-basin and oriental sub-basin).
-The existence of two sub-basins and a structural higher area between them, allied with the expansion and retraction of the seas allowed a variation of sedimentation environments, ranging between the external platform and internal platform, and continental, hemipelagic, or deep marine environments (Terrinha et al., 2013).
-It was this variance in deposition environments that created the variety of Paleozoic sedimentary facies, with moments of more or less homogeneity throughout this period.
-The Algarve basin is of essential importance for the existence of chert in the Algarve region since it is in the Paleozoic sediments where chert nodules can be found.
-Paleozoic sedimentation in the basin starts in the Triassic, although the identification of chert nodules or beds is limited to the Lower to Upper Jurassic.
-It is at the start of the Lower Jurassic, during the Lower Pliesbachian (also named Carixian) that the Algarve basin started to separate into two sub-basins until the Callovian (Middle Jurassic) when an uplift episode allowed uniform sedimentation of the basin (Rocha, 1976; Terrinha et al., 2013).
-During the Lower Pliensbachian, the sediments in the western sub-basin can be described as marine, of the external platform, while the sediments of the eastern sub-basin are of an internal platform.
-The transition between the Lower Jurassic to the Middle Jurassic shows a hiatus, as there are no known sediments from the Toarcian to the Aaelinian in the basin, the sedimentation restarting only after the latter age.
-During the Upper Jurassic, the lithofacies variation became more prominent, especially during the Upper Oxfordian-Lower Kimmeridgian, with a certain uniformity of the lithofacies in all the sub-basins during the Upper Kimmeridgian (Terrinha et al., 2013).
+It extends for \~130 km E-W and \~50 km N-S and is characterized by a variety of geomorphic sub-regions and geological units, that make this region a complex territory (\@ref(fig:geomap)).
+On the north sector of the Algarve, the Serra Algarvia is characterized by a mountainous range with a dense hydrographic network, which separates the Algarve from Alentejo.
+On the south sector, the Litoral is characterized by a flatter, long strip of land, that extends through all of the coastal strip of the Algarve.
+The Barrocal is nested between the other sub-regions and has a more moderate relief, characterized by carbonated Jurassic formations and important sub-terranean water circulation [@gago2007].
+
+```{r geomap, echo=FALSE, fig.cap="Geological map of the Algarve region, bordered north by the Alentejo region and east by Spain.
+The map represents the several geological layers and formations, as well geomorphic sub-regions.
+The Jurassic layers are colored in green.
+Metadata was obtained through the SNIAmb (Sistema National de Informação de Ambiente)", out.width='100%'}
+
+knitr::include_graphics("../figures/geology_map.png")
+
+```
+
+Geologically, the Algarve is composed of two main geological units: the South Portuguese Zone (SPZ) and the Algarve basin.
+The SPZ is located in the north sector of the Algarve, extending up to Alentejo [@fernandes_new_2012].
+Its main lithologies are schist, greywacke and quartzite [@fernandes_new_2012; @pereira2016].
+The SPZ is overlain unconformably by the Mesozoic sedimentary rocks of the Algarve basin [@fernandes_new_2012].
+The basin corresponds to the Mesozoic-Cenozoic sediments that outcrop south of the Algarve, from the westernmost to the easternmost point of the region, and it is associated with the opening of the central Atlantic Ocean and with the eventual oceanic crust formation in the western part of the Tethys sea, between the Algarve and North Africa [@terrinha_bacia_2013].
+Paleozoic sedimentation of the basin started in the Triassic and continued thereon.
+In the Lower Jurassic (Lower Pliensbachian, also regionally known as Carixian) the basin was divided into two sub-basins -- western and oriental sub-basins [@rocha1976; @terrinha_bacia_2013].
+The existence of the two sub-basins and the expansion and retraction of the sea created a variety of sedimentation environments, such as external and internal platforms, continental, hemipelagic and deep marine [@terrinha_bacia_2013], as well as moments of sedimentation hiatus.
+This variability in deposition environments created a variety of Paleozoic sedimentary facies, with moments of more or less homogeneity throughout this period.
+For example, during the Lower Pliensbachian, in the Lower Jurassic, the sediments in the western sub-basin can be described as marine of external platform, while the sediments of the eastern sub-basin can be described as marine of internal platform.
+During the Upper Jurassic however, the basin is marked by a moment of prominent lithofacies variation, followed by a moment of uniformity in both sub-basins [@terrinha_bacia_2013].
+Understanding the Algarve basin is key for raw material studies in the Algarve, especially when studying chert, since it is in the basin, more specifically in the Paleozoic sediments, where chert primarily outcrops in the region.
 
@@
 
-In general, the presence of chert in the Algarve region and be attributed to limestone or dolomitic limestone layers.
-They appear with relative frequency, and often as nodules of small to medium dimensions, although thin beds of chert can also be identified in some areas.
-As mentioned above, outcrops with chert nodules in the Algarve region can be found in the Paleozoic sediments from the Algarve Basin, more specifically in the Jurassic sediments.
-Chert nodules can be found within Lower, Middle and Upper Jurassic layers, although restricted to a limited number of layers within the stratigraphy and not present in every geological cut.
-
-Lower Jurassic geological layers with chert nodules can be mostly found in the Carixian-age sediments (also known as Pliensbachian) from the Sagres region, often outcropping in those places where more layers are exposed, such as beach-generated cliffs and associated deposits.
-These include the chert nodules and beds from Cabo de S.
-Vicente to Praia do Belixe, but also the small outcrop inland, in Ferrel.
-The chert nodules occur within the limestone, dolomitic limestone, and marly limestone.
-Further east one layer with chert nodules was also identified but attributed to the Sinemurian-Toarcian, in a formation identified as “Limestones with Paleodasucladus and limestones with chert nodules”.
-Described only in the Geologic Map at a 1 200 000 scale, however, these chert nodules were characterized as micronodules [@oliveira_carta_1992].
-
-Middle Jurassic geological layers with chert nodules are only found east of Sagres, in the Albufeira, Faro, and Tavira regions.
-These layers were attributed to the Malhão Formation, from the Aalenian-Bajocian age.
-In general, this formation was described as carbonated, marine, and thick, and chert can be found in two different layers: monogenic conglomerates with micritic limestone intercalations with chert beds and nodules; microcrystalline limestones with chert nodules (pel-biomicrite).
-Another chert outcrop was identified in the Geologic Map at a 1 200 000 scale, attributed to the Bajocian-Batonian age and named “Alagoa conglomerates”, with partially dolomitic limestones and chert nodules.
-
-Finally, Upper Jurassic sediments have been identified also in the Albufeira, Faro, and Tavira area.
-The Jordana Formation was attributed to the Kimmeridgian/Kimmeridgian-Portlandian and is characterized by dark-gray limestones, with frequent secondary silifications with abundant fossil fragments.
-In the Tavira area, these layers were instead attributed to the Oxfordian-Kimmeridgian age.
+The presence of chert is commonly associated with carbonates in limestone and dolomite formations.
+This is explained by characteristics (such as the presence of water or specific temperatures and pH) that are ideal for both the formation of limestone and the precipitation of silica [@luedtke1992].
+The pelagic and marine environments of the Algarve basin during the Paleozoic gathered those such ideal characteristics, allowing the sedimentation and precipitation of limestone and silica, creating limestone layers with chert nodules.
+The existence of two basins with different sedimentation environments also shows potential for the precipitation of different types of chert throughout the basin and their differentiation.
+For example, during sedimentation, skeletal grains of fossils may be preserved.
+Many of these fossils are restricted to specific sedimentation environments and time intervals [@flügel2010], which may allow the identification of chert outcrops through the fossil content.
+The basin and sub-basins thus show potential for the existence of different geological formations with different chert types and their study.
+Previous works, both geological and archaeological, confirm that chert is present in the Algarve in the Jurassic limestone or dolomitic limestone layers of the Algarve basin.
+This means that chert outcrops can be identified in the central/south sector of the region, from west to east.
+Variability in chert availability, as well as chert types, is expected, considering that during the sedimentation process, the Algarve basin was already sub-divided and in constant environmental change.
+Due to this, several formations with chert nodules can be identified in the Algarve, corresponding to different sub-periods of the Jurassic.
+In the western sector of the Algarve, chert can be found in the Lower Jurassic limestone or dolomitic limestone layers (from the regional Carixian), often visible in areas where the layers are exposed, such as beach-generated cliffs and associated deposits.
+This includes outcrops such as Cabo de S.
+Vicente (CSV) and Praia do Belixe (PBLX).
+These sediments and chert nodules are also available inland, albeit more scarcely, as is the case of the small outcrop named Ferrel, 3 km from the current coastline.
+Lower Jurassic chert-bearing outcrops are barely existent in the center/east sector of the Algarve, with one single formation with micro-nodules identified in geological works [@oliveira1992].
+Middle Jurassic geological layers with chert nodules are only found in the center/east sector of the Algarve in a geological formation called the Malhão Formation.
+The formation can be described as carbonated, from a marine sedimentation environment .
+Chert in this formation has been identified in two distinct layers: conglomerates with micritic limestone intercalations with chert beds and nodules, characterized by the presence of sponge spicules and radiolarians [@manuppela1987]; microcrystalline limestones with chert nodules characterized by the presence of silicified malacofauna and silicified corals [@manuppella2007].
+Finally, Upper Jurassic sediments with chert nodules have also been mostly identified in the center/east sector of the Algarve, attributed to the Jordana Formation.
+This formation is characterized by dark-gray limestones, with frequent secondary silicifications with abundant fossil fragments [@manuppela1987; @rocha1989; @manuppella2007].
+One single outcrop from the Upper Jurassic has been identified in the western sector of the Algarve basin, located at Praia da Mareta.
 
@@
 
-To identify the chert outcrops present in southern Portugal and understand the chert’s characteristics, a macroscopic and petrographic approach was applied to the study of geological samples.
-A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials.
-This is because different methods have their inherent limitations.
-By combining these two different methodologies the goal was to create several layers of analysis, which may then be further pursued or added new layers (Brandl, 2016).
-Several other similar methodologies and approaches have been applied in other regions (references).
-However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, problematics, and the characteristics of the types of cherts in question.
-Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region, the two methodologies were chosen for the study.
-Geological samples were obtained through fieldwork, from August of 2021 (west Algarve) and April to June 2022 (east Algarve), and included visits to previously known outcrops and prospection of areas identified in geological maps.
-The first step of fieldwork included the review of previously known research, including geological maps, archaeological prospection works, and raw material studies (Pereira, Verissimo 2004, 2005).
-These were followed as a reference guide for recovering samples.
-Whenever necessary, samples from previous prospection works were consulted, from the LusoLit lithoteque from ICArEHB at the University of Algarve (Pereira) and the thin sections at the University of Évora.
-Whenever coordinates or specific locations for known outcrops were available, those were visited, and then the surrounding area was prospected to understand the extension of the outcrops and possible secondary deposition outcrops nearby.
-In all other situations, several locations with the potential to find outcrops were selected from polygons where formations with chert were known to the outcrop.
+To locate and characterize chert outcrops in southern Portugal and understand chert's characteristics, a macroscopic and petrographic approach was applied to the study of geological samples which were collected through fieldwork.
+A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials, especially since different methods have their inherent limitations.
+By combining these two different methodologies the goal was to create several layers of analysis, to which new layers may be added if necessary [@brandl2016].
+Several other similar methodologies and approaches have been applied in other regions [i.e., @garcía-simón2016; @terradas2017; @tomasso2019; @desoler2020].
+However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, the problematics, and the characteristics of the types of cherts in question [@luedtke1992].
+Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region [@ribeiro2005], the two methodologies were chosen for the study.
+The geological samples used in this study were obtained during fieldwork, between August 2021 and June 2022.
+The locations prospected were chosen after reviewing previously known research, which included preliminary raw materials studies in the region [@verissimo2004; @pereira2016], geological scientific papers and theses focusing on the Algarve basin and concerning chert outcrops [@rocha1976; @marques1983; @ramalho1985; @ribeiro2005], and geological maps, which signaled the presence of chert nodules within the outcrops[@rocha1979; @rocha1983; @rocha1989; @oliveira1984; @oliveira1992; @manuppela1987; @manuppella2007].
+Unpublished data and coordinates for unsurveyed locations with potential for chert-bearing outcrops gathered during the organization of the LusoLit lithotheque were also checked.
+Whenever coordinates or specific locations for known outcrops were available, these were directly visited and the surrounding area was surveyed to understand the extension of the outcrops and to locate possible secondary deposition outcrops nearby.
+When no specific locations were described (for example, in geological maps) several areas with more potential to find chert outcrops within one larger formation were surveyed.
 Samples were collected whenever possible, focusing on both primary and secondary outcrops.
-When the outcrops showed macroscopic differences between the cherts, several samples were collected, to obtain as much variability as possible.
-All samples were georeferenced and registered with resources to an android app which allows accurate coordinates to be associated with photos and CSV data regarding the outcrop characteristics and conditions (i.e.
+When chert nodules within one single outcrop showed macroscopic differences (such as differences in the color, texture, translucency, or cortex), samples of each different nodule were collected, to cover all chert variability within the outcrop.
+This variability was also recorded using a database (to distinguish between homogeneous or heterogeneous chert nodules within the outcrop) and through photography (\@ref(fig:fieldphoto)).
+All samples were registered with resource to a free android app (Archaeosurvey) which was designed for archaeological surveys, and records site location and characteristics [@cascalheira2017a], further adapted for raw material source surveys [@abrunhosa2017].
+The version of the software used for fieldwork is an adaption of the latter apps and records data related to outcrop characteristics and conditions (i.e.
 abundance, visibility, access, geomorphology, chert morphology, and conditions).
-All data related to the app and dataset for the variables recorded can be found in the supplementary materials.
-Individual ID tags were associated with each sample, which was washed and labeled in the laboratory.
-All geological samples  were analyzed macroscopically following a pre-established dataset.
-The variables were defined based on specialized literature and the dataset with the variables can be found in the Supplementary materials.
-The goal of this analysis was to describe the cherts and their enclosing rocks in a detailed and objective manner.
+All data related to the app and fieldwork can be found in the Supplementary Online Materials (SOM).
+Individual IDs were associated with each sample, which includes sequential numbers (based on recovery order, i.e., SP10) and outcrop code (i.e., PdA).
+
+```{r fieldphoto, fig.cap="Sample SP65_MALH from the Malhão outcrop, collected during fieldwork.
+Left – chert samples collected and nodule still embedded in the parent rock.
+All chert macroscopic variability was recoded through photography upon sample collection.
+Right – general photo of the outcrop.
+The outcrops and points of reference were recorded for contextualization and easiness of future access.", out.width='100%'}
+
+knitr::include_graphics("../figures/fieldwork.png")
+
+```
+
+A two-step approach was applied to characterize the geological samples.
+First, the samples were analyzed macroscopically following a pre-established dataset.
+The variables were defined based on specialized literature [@luedtke1992; @crandell2005; @bressy2002], and the dataset with variable description as well as the sample characterizations can be found in the SOM.
+A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with resource to a Nikon SMZ25 stereomicroscope, focusing primarily on inclusions and fossil content.
+Despite several caveats, especially related to the subjectivity and lack of quantitative variables [@bustillo2009], a macroscopic approach is currently still frequently used in chert raw material studies.
 For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages.
 Macroscopic analyses have the advantage of being less costly and easy to apply.
-Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops, and formations is essential to apply the methodology with archaeological materials and establish comparisons (Bustillo et al 2009 ).
-A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with the resource of a Nikon SMZ25 stereomicroscope, focusing especially on aspects like inclusions and fossil content.
-Thin sections were produced for geological samples of all formations, focusing on obtaining petrographic data which reflected the variability observed at a macroscopic level.
-In total, 29 thin sections  were produced (thin-section database info), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve, recovered from the 2021 fieldwork; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve, recovered from the 2022 fieldwork; 3) 1 thin-section of a geological sample recovered from previous works (<2021), which was not identified in the latter prospections.
-Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced, to characterize the chert from the most visited locations.
-Information about the thin sections, including the laboratories in which they were made, can be consulted in the Supplementary materials.
-All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in the Supplementary materials).
-Three geological samples from western Algarve (previously broken for thin-section production) were also chosen for X-ray diffraction analysis, which was made at the GeoSciences CSIC institute in Barcelona, using a Bruker D8-A25 with a Cu tube and ultra-fast PSD detector.
-All descriptions (macroscopic, petrographic, and XRD data) and accompanying photographs will be available online on a database, and downloadable as PDF files.
-The paper was written with resource to R and R Studio and can be found open-source at the **link**.
+Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops and formations is essential for comparative studies between geological samples and archaeological assemblages and obtaining preliminary results.
+By then combining the macroscopic analysis with a petrographic analysis, other studies have shown that the subjective component of this approach can be minimized [@bustillo2009].
+The second phase of the study focused on the petrographic analysis of the geological samples.
+Thin sections were produced from geological samples of all formations, focusing on obtaining petrographic data that reflected the macroscopic variability.
+In total, 30 thin sections were produced (\@ref(tab:petrolist)), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve; 3) 1 thin-section of a geological sample recovered from previous works, which was not identified during our survey.
+Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced.
+To compare with the thin sections from this study, other thin sections from previous studies of Jurassic outcrops from western Algarve [@ribeiro2005] were also consulted.
+All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in the SOM).
+
+```{r petrolist, echo=FALSE}
+
+petrolist <- read_csv("../data/raw_data/petro_sample.csv")
+knitr::kable(petrolist, "simple", caption = "List of geological samples chosen for the petrographical study. UB - University of Barcelona; TSL - Thin Section Lab.")
+
+```
+
+All sample descriptions (macroscopic and petrographic) and accompanying photographs, as well as photographs and data about the outcrops are also available on a database dedicated to the LusoLit lithotheque , which can be accessed online at <https://lusolit.icarehb.com/>.
+This database will continue to be updated with the other existing samples, non-chert raw materials and future analyses.
+The complete R code used for all the analysis and visualizations contained in this paper is available at our online research compendium (DOI: 10.17605/OSF.IO/FP7TA).
+To produce those files, we followed the procedures described by Marwick et al.
+(2017) for the creation of research compendiums to enhance the reproducibility of research.
+The files provided contain all the raw data used in our analysis as well as a custom R project [@wickham2015] holding the code to produce all tables and figures.
+To enable maximum reuse, code is released under the MIT license, data as CC‐0, and figures as CC‐BY (for more information, see [@Marwick2017].
 
@@
 
-Eighteen outcrops (primary and secondary) were revisited or identified in the Algarve region, nine in the westernmost territory, and nine to the east (between Loulé and Tavira).
-From these, 70 samples were recovered and analyzed.
-Some of these samples were recovered as isolated finds or in secondary settings.
+Eighteen outcrops (primary and secondary) were revisited or newly identified in the Algarve region.
+Nine are located in the westernmost territory and nine to the east (\@ref(fig:sourcesmap)).
+From these, 57 samples were recovered and analyzed, of which 19 are isolated finds or in secondary settings (\@ref(tab:outcropdata)).
 
-On the westernmost part of the Algarve, there are mainly cherts from two different formations: Lower Jurassic and Upper Jurassic.
-The latter can be found in a single known outcrop - Praia da Mareta.
-Lower Jurassic outcrops are more common and have been better studied (ribeiro_evolucao_2005).
-These outcrops are heterogeneous, showing different characteristics and chert colors.
+```{r sourcesmap, fig.cap="Map of southern Portugal (Algarve), with geological samples recovered during this study’s fieldwork.
+Colors represent the different geological ages, based on the formations where the cherts were found.
+N/A values represent samples (secondary sources or isolated finds) to which a geological age or formation could not be attributed.", out.width='100%'}
 
-The Lower Jurassic cherts can be grouped in three main macroscopic types: yellow and/or red type; grey/brown type; yellow and red with fossils type.
-These macroscopic types not only show color differences but also in the fossil content, visible to the naked eye and stereomicroscope.
+knitr::include_graphics("../figures/sources_map.png")
+
+```
+
+```{r outcropdata, echo=FALSE}
+
+outcrop_db <- read_csv("../data/raw_data/outcrop_short.csv")
+
+knitr::kable(outcrop_db, "simple", caption = "Outcrop and nodule information for all recovered samples.
+The age was defined based on the location of the outcrops, taking into consideration previous research work and geological maps.
+Samples which were isolated or in untrackable secondary deposition settings do not have a defined geological age.")
+
+```
+
+## Western Algarve
+
+On the westernmost areas of the Algarve, there are mainly cherts from two different formations: Lower Jurassic and Upper Jurassic.
+The latter can be found in primary deposition in a single known outcrop - Praia da Mareta -- or nearby, in secondary deposition settings (\@ref(fig:sourcesmap)).
+Lower Jurassic outcrops are more common and, for that reason, have been more studied [@ribeiro2005].
+These outcrops are heterogeneous, showing different geological characteristics and chert colors (\@ref(fig:ljmacro)).
+The Lower Jurassic cherts can be grouped into three main macroscopic types based on color (individual Munsell Color Chart codes can be found in the macroscopic description analysis table) and presence of fossil content: multicolored, yellow, red or purple type (MC); single grey/brown type (SGB); multicolored, yellow, red or purple with fossils type (MCF).
 The first two types are present in all outcrops.
 They are mainly characterized by dull to medium luster and opaque translucency, although some samples were sub-translucent.
-The feel ranges between smooth and semi-smooth, although many of the Belixe cherts have a rough feel.
-In the Yellow/Red cherts, fossil content is present but visible only as white, red, or yellow speckling.
-The Grey/Brown cherts show little fossil content, barely visible with the stereomicroscope.
-The Yellow/Red with fossils cherts show a large number of big fossils, which are easily seen by the naked eye and can be identified under the stereomicroscope.
-According to the microscopic approach, the Lower Jurassic cherts of Western Algarve are composed mainly of microcrystalline quartz, with textures that range mostly from wackestone to packstone.
-In more than 50% of the samples, no fossil can be identified, as all fossils, albeit common to very frequent, are poorly preserved, filled with chalcedony or quartz, and without any identifiable morphology.
-Whenever identifiable, fossils present in the sample are Echinoderms, Radiolarians, Sponge spicules, and a Bivalve shell.
+The feel ranges between smooth and semi-smooth, although many of the Belixe cherts are distinctly rough to the touch .
+In the MC cherts, fossil content is present but visible only as white, red, or yellow speckling.
+The SGB cherts show little fossil content, barely visible with the stereomicroscope.
 
-Despite the similar characteristics between these cherts, independent of color or geographical location, the outcrops are heterogeneous and show varying characteristics, which may be of importance to distinguish between chert sources within the Lower Jurassic formation.
+```{r ljmacro, fig.cap="View of the macroscopic variability of the Lower Jurassic cherts from western Algarve.
+A- SP34_PdA ; B- Sample SP14_PBX; C- Sample SP16_PBX; D- Sample SP30_FZF; E- SP9_PdA; F- Sample SP34_PdA.", out.width='100%'}
+
+knitr::include_graphics("../figures/macro_lowerJ_east.png")
+
+```
+
+The MCF show a large quantity of larger fossils (\~1000 µm), which are easily seen by the naked eye and can be identified under the microscope.
+Petrographically, the Lower Jurassic cherts of western Algarve are composed mainly of microcrystalline quartz, with textures that range mostly from wackestone to packstone.
+In more than 50% of the samples, no fossils can be identified, as all fossils, albeit common to very frequent, are poorly preserved, filled with chalcedony or quartz, and without any identifiable morphology.
+Whenever identifiable, fossils present in the sample are Echinoderms, Radiolarians, Sponge spicules, and bivalve shells (\@ref(fig:ljwest)).
+
+```{r ljwest, fig.cap="Column A – Sample SP7_CSV.
+Above: Macroscopic view with a stereomicroscope; Center and below: Microscopic view of a thin section (XPL and PPL respectively).
+An Echinoid spine, replaced by quartz but preserving the fossil’s original structure is present in the center of the images.
+Column B – Sample SP34_PdA.
+Above: Macroscopic view with a stereomicroscope; Center and below: Microscopic view of a thin section (XPL and PPL respectively).
+Includes the detail of a Bivalve shell replaced by at least two generations of quartz: microcrystalline quartz at the edges and macrocrystalline quartz inside of the shell.", out.width='100%'}
+
+knitr::include_graphics("../figures/lowerJ_west.png")
+
+```
+
+Despite the similar characteristics between these cherts, the outcrops are heterogeneous and show varying characteristics between them, which may be of importance to distinguish between chert sources within the Lower Jurassic formation.
 These outcrops have been divided into four groups, following the available literature: Cabo de S.
 Vicente (including Cabo de S.
-Vicente and Aspa); Foz dos Fornos; Ponta dos Altos; Praia do Belixe (which includes Belixe Sul); and Ferrel.
-
+Vicente and Aspa), Foz dos Fornos, Ponta dos Altos, Praia do Belixe (which includes Belixe Sul), and Ferrel.
 The Cabo de S.
@@
 The banks seem to be mainly dolomite or dolomitic limestones.
-The process of dolomitization seems to have affected the chert nodules, as they often present different levels of silicification from the peripheral areas of the nodule to the interior (which is more silicified), which also affects the knapping quality.
-The nodules vary in size, ranging from small 4 cm in diameter circular nodules to bed-like groups of nodules of ~20 cm in width.
+The process of dolomitization seems to have affected the chert nodules, as they often present different levels of silicification from the peripheral areas of the nodule to the interior, which also affects the size and feel of the grain .
+In this case, the peripheral areas of the chert nodules are more dolomitized, with visible grain and distinctively rough to the touch, while the interior areas are more silicified and conversely finer and smoother.
+The nodules vary in size, ranging from small 4 cm in diameter circular nodules to bed-like groups of nodules with \~20 cm in width.
 At the Aspa outcrops, the nodules are less frequent and smaller.
@@
 Foz dos Fornos (FZF) and Ponta dos Altos (PdA) show similarities to the CSV outcrops.
-The nodules are visible in several banks of dolomite, dolomitic limestone, and limestone, partially covered by soil.
+The nodules are visible in several banks of dolomite, dolomitic limestone, and limestone, partially covered by soil and sand.
 The nodules can be circular, around 5 cm in diameter, or wide nearly 20 cm in width.
@@
 Between the FZF chert and the PdA, the main differences seem to be the cortex and parent rock, which show differing reactions to hydrochloric acid, the first being dolomite or dolomitic limestone, and the second being mostly limestone, with some degree of dolomitization in certain areas.
-
-Praia do Belixe (PBLX) is characterized by the abundance of chert nodules throughout the dolomite layers of the cliff area.
+Praia do Belixe (PBX) is characterized by the abundance of chert nodules throughout the dolomite layers of the cliff area.
 They are visible in certain areas of the cliff and within the rock shelters.
-The nodules can be small, around 5 cm in diameter, sometimes reaching more than ~30 cm in width, or bedded, as chert layers between the dolomite layers.
-The cherts show varying degrees of dolomitization and are mostly characterized by coarse to semi-smooth feel, dull luster, and a medium to low knapping quality.
-Unlike the other outcrops, no chert nodule fragments were found close to the cliffs, and samples could only be recovered directly from the embedded nodules in the cliff walls.
+The nodules can be small, around 5 cm in diameter, sometimes reaching more than \~30 cm in width, or bedded, as chert layers between the dolomite layers.
+The cherts show varying degrees of dolomitization and are mostly characterized by a coarse to semi-smooth feel and dull luster, often showing fractures and alterations.
+Unlike the other outcrops, no chert nodule fragments were found close to the cliffs, and the samples could only be recovered directly from the embedded nodules in the cliff walls.
 Nodules scattered on the floor were only located at Belixe Sul (BLS), a primary outcrop nearly destroyed located on a field, north of the beach area.
-The chert in this outcrop showed no differences from PBLX, aside from the size of the nodules, which were smaller and often showed signs of post-depositional alterations.
-
-The third location for chert has been previously identified north of BLS.
-Belixe Norte (BLN) is located on a dirt road and unused agriculture field.
+The chert in this outcrop showed no differences from PBX, aside from the size of the nodules, which were smaller and often showed signs of post-depositional alterations.
+A third location for chert has been previously identified north of BLS.
+Belixe Norte (BLN) is located on a dirt road and an unused agriculture field.
 Several chert fragments were collected in this location.
@@
 No larger nodules or outcrop were identified in this location.
-The samples recovered from the location also seem to corroborate that BLN should not be considered an outcrop, as they do not match the local cherts and rather, resemble most of the samples recovered from the Eastern Algarve.
+The samples recovered from the location also seem to corroborate that BLN should not be considered an outcrop, as they do not match the local cherts and rather, resemble most of the samples recovered from eastern Algarve.
 Ferrel, unlike the other outcrops, is located inland and away from the coast.
@@
 All surface fragments and nodules were small, with around 2 to 3 cm of width which may be explained by the state of the outcrop.
-Unlike the outcrops from the Lower Jurassic of the western Algarve, this region of the Algarve has only one identified outcrop for Upper Jurassic cherts.
+Contrasting with diversity of the Lower Jurassic there is only one identified outcrop for Upper Jurassic cherts in western Algarve.
 These are located at Praia da Mareta (MAR) and abundant, or in a secondary deposition at Ponta da Atalaia (PtA).
-At Praia da Mareta the nodules are only easily accessible on the beach, where large chunks of the cliff (~1 m in diameter) are transported by the waves.
+At Praia da Mareta the nodules are only easily accessible on the beach, where large boulders falling off the cliff (\~1 m in diameter) are transported by the waves.
 Several chert nodules of different sizes can be found in the parent rock washed ashore, ranging between 2 cm to 20 cm in diameter.
@@
 The only identifiable difference is the presence of calcispheres.
-All samples from the Mareta outcrops seen under the petrographic microscope showed the presence of abundant calcispheres , which is not always apparent with the stereomicroscope.
-Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic, which was uncommon and scattered at the top of the cliffs by the beach.
+All samples from the Mareta outcrops and PtA seen under the petrographic microscope showed the presence of abundant calcispheres (\@ref(fig:ujwest)), which is not always apparent with the stereomicroscope.
+Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic (\@ref(fig:ujwest)), which were uncommon and scattered at the top of the cliffs by the beach.
+
+```{r ujwest, fig.cap="Column A – Sample SP36_PtA.
+Above: Macroscopic view with a stereomicroscope; Center and below: Microscopic view of a thin section (XPL and PPL respectively).
+Several unidentifiable fossils can be seen in the photo, along with the calcispheres.
+Column B – Sample SP39_AND.
+Above: Macroscopic view with a stereomicroscope; Center and below: Microscopic view of a thin section (XPL and PPL respectively).
+A small amount of calcispheres is present in the image, along with few unidentifiable fossils replaced by quartz/chalcedony.", out.width='100%'}
+
+knitr::include_graphics("../figures/upperJ_west.png")
+
+```
+
+## Eastern Algarve
 
 On the eastern part of the Algarve, chert-bearing known formations are from the Middle to Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively.
-
-The Malhão formation chert (dating to the Middle Jurassic) was identified in three outcrops, in the Faro parish and the Tavira parish.
+The Malhão formation chert (dating to the Middle Jurassic) was identified in three outcrops.
 Whenever in a primary outcrop, this chert was homogeneous.
@@
 In these outcrops, the nodule frequency varied from common to abundant.
-The nodules are roundish, ranging between 3 to 5 cm of maximum width.
+The nodules are roundish, ranging between 3 to 5 cm in maximum width.
 In all cases, access to the outcrops was easy.
@@
 Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified.
-The petrographic analysis shows for the Malhão cherts from Casal da Colina (primary and secondary outcrops) have high amounts of dolomite within the chert.
-All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all poorly preserved and replaced by chalcedony or quartz).
-These fossils are Sponge spicules, Radiolarians, Ostracods, Tentaculites (also previously identified by – ), Echinoderms, and Calcispheres.
+The petrographic analysis shows for the Malhão cherts from Casal da Colina (primary and secondary outcrops) high amounts of dolomite within the chert.
+All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all are poorly preserved and replaced by chalcedony or quartz).
+These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (Fig.
+\*\*).
 
-The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve, above Moncarapacho in the Olhão parish.
+```{r mjplate, fig.cap="Macroscopic and microscopic views of chert samples from the Middle Jurassic Malhão outcrop.
+A – Sample SP50_CdC, macroscopic view with a stereomicroscope.
+B – Sample SP62_OLV, macroscopic view with a stereomicroscope.
+C – Microscopic view of SP50_CdC, XPL (left) and PPL (right).
+Detail of a fossil (possibly an Ostracod), replaced by two generations of chalcedony (1st generation in the outer edges and 2nd generation replacing the inside).
+D - Microscopic view of SP54_GUI, XPL (left) and PPL (right).
+General view of the thin section.
+Several fossil ghosts can be seen.
+Despite the poor preservation, it may be possible to identify a few fossils based on the size and morphology: 1) calcispheres  or recrystallized radiolarians; 2) monaxon spicules pointed at one end.", out.width='100%'}
+
+knitr::include_graphics("../figures/middleJ_east.png")
+
+```
+
+The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (Fig: ).
 Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock.
@@
 Smaller nodules broken from the parent rock were identified near the primary source in a field.
-Whenever embedded in the parent rock, the nodules varied in size (~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock.
+Whenever embedded in the parent rock, the nodules varied in size (\~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock.
 The cherts show little macroscopic variability between nodule and outcrop.
 They are grey/brown (with visible yellow inclusions).
-Within nodule, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas.
+Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas.
 Some of the nodules also show the variability of translucency, with translucent areas, with a very fine grain, and little presence of visible inclusions.
@@
 They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals.
-Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shells, Sponge spicules, Ostracod, Echinoderms, and Gastropod.
-Perhaps add a paragraph about the sources which were marked on the geological map but could not be located.
+Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (Fig.
+\*\*).
+
+```{r ujplate, fig.cap="Macroscopic and microscopic views of chert samples from the Upper Jurassic Jordana outcrop.
+A – Sample SP58_JOR, macroscopic view with a stereomicroscope.
+B – Sample SP59_JOR, macroscopic view with a stereomicroscope.
+C – Microscopic view of SP56_JOR, XPL (left) and PPL (right).
+General view of the calcispheres in an area of the chert characterized by a large concentration of oxide patina and opaques.
+D - Microscopic view of SP58_JOR, XPL (left) and PPL (right).
+General view of the thin section.
+A large bivalve shell fragment is visible at the top.", out.width='100%'}
+
+knitr::include_graphics("../figures/upperJ_east.png")
+
+```
+
+## Other outcrops
+
+It is important to note that the aforementioned chert geological samples represent the chert variability of the identified outcrops.
+However, a small number of outcrops described in regional geological maps were not identified or were not accessible.
+This includes four outcrops: 1) a Lower Jurassic outcrop with chert micronodules which was identified through a geological profile (Oliveira, 1992); 2) a Middle Jurassic outcrop located at the easternmost section of the Algarve with partially dolomitized clasts and chert nodules (Oliveira, 1992); 3) several unprospected locations from previous geoarchaeological works, all related to the Peral Anticline and reaching from central to eastern Algarve; 4) a possible chert exploitation archaeological site located on top of a chert source (Monte do Cerro), in eastern Algarve which was not identified and is possibly inaccessible (Bicho et al., 2003).
+All of these unidentifiable outcrops are located in the central or eastern portion of the Algarve.
+This may be related to the frequent landscape changes occurring due to agriculture or the population of previously uninhabited areas, which seldom occurs in western Algarve by the cliff areas, where most outcrops are located.
 
@@
 
-The prospection works and analyses of the collected geological samples show that the south of Portugal has a high potential for raw material studies, especially regarding chert.
+The survey work and analyses of the collected geological samples show that the south of Portugal has a high potential for raw material studies, especially regarding chert.
 The presence of chert-bearing outcrops in the westernmost part of the Algarve, center, and east would provide several possibilities for sourcing and procurement whenever groups moved throughout the territory.
-This is further important when we consider the geology of this territory.
+This is further important when we consider the geology of this region.
 The geology of the Algarve itself may have played an important part in how groups procured their raw materials, specifically, their chert, a task that has been identified as essential for hunter-gatherer groups.
-To the south, communities would only have access to chert-bearing outcrops up to the coast.
+To the south, communities would only have access to chert-bearing outcrops down to the coast.
 To the north, the mountain range would not only have provided no chert nodules but may have also hampered the movement of populations, forcing groups to move east and west instead of north or south.
 This movement may have facilitated the gathering of cherts from different formations within the Algarve, posteriorly then brought into the sites.
-Especially for Middle and Upper Paleolithic occupations, understanding the sources of chert in the Algarve may provide data about where in the territory they were sourcing their raw materials, and how they were using the territory when their movement was constrained by the Peninsula’s natural barriers.
-This question remains unanswered, but raw material data and comparative studies between archaeological assemblages and geological samples may provide the necessary results to answer it.
-
+Especially for Middle and Upper Paleolithic occupations, understanding the sources of chert in the Algarve may provide data about where in the territory these groups were sourcing their raw materials, and how they were using the territory having in consideration the region's natural barriers and consequent distribution of resources.
+Although this topic remains unexplored, this study stands as a further step to tackle these questions in the Algarve, as it may provide the necessary basis for comparative studies with archaeological assemblages.
 However, tracking these movements and procurement patterns is only possible if the cherts from the different formations and outcrops can be traced back to their sources.
@@
 Within formations, however, there are no discernable differences, both at a macroscopic and petrographic level, as these do not seem to be useful to distinguish between outcrops.
-That is most obvious in the Lower Jurassic outcrops of west Algarve.
+This is most obvious on the Lower Jurassic outcrops of western Algarve.
 The identified chert groups, which varied mostly in color and fossil content, are present in several outcrops.
-In this region, the variables which may be better used to understand which outcrops were visited may be the quality and size of the nodules.
-The latter, for example, is an important variable in the Belixe outcrops, which show the largest volumes of rock, even if the knapping quality is worse than some other available, smaller nodules.
-Size may be used in conjunction with other technological data, to understand whether different nodules were being explored differently based on their size, or their procurement was being preferred over other smaller nodules in possibly closer outcrops in the region.
-Size may also be an important variable to distinguish between nodules collected in primary or secondary outcrops.
-Whenever found in secondary outcrops (without the presence of parent rock and with rounded and/or altered cortex) the nodules were small and altered.
-Although in other regions, secondary outcrops such as river beds, for example, may be of extreme importance to raw material procurement, it seems that for the regional cherts, these would be procured in primary outcrops.
-
-The Upper Jurassic nodules of the western Algarve also show larger volumes than those from the Lower Jurassic.
-This, together with the translucency, which is often characterized by sub-translucent fabrics, may be a good macroscopic indicator of the chert’s provenance.
+In this region, the variables which may be better used to understand which outcrops were visited may be the quality (differing levels of dolomitization, presence/absence of fractures or even size of grain) and size of the nodules.
+The latter, for example, is an important variable in the Belixe outcrops, which show the largest volumes of rock, even if the chert's quality is worse than some other available, smaller nodules.
+Size may be used in conjunction with other technological data, to understand whether different nodules were being explored differently based on their size, or their procurement was being preferred in relation to other smaller nodules in possibly closer outcrops in the region.
+The Upper Jurassic nodules of western Algarve also show larger volumes than those from the Lower Jurassic.
+Translucency also seems to be a good macroscopic indicator to distinguish between western Algarve Lower and Upper Jurassic cherts, since the latter are characterized as sub-translucent.
 However, for a reliable distinction between the Lower Jurassic and the Upper Jurassic cherts, petrographic analyses and the identification of calcispheres may be necessary.
-The differences identified between the cherts of the various formations can be seen both at a macroscopic and petrographic level.
-Given the formation setting, petrographically, all the cherts from the Algarve are fairly homogeneous - marine origin, in limestone or dolomitic limestone rocks, all formed during the Jurassic.
+The differences identified among the cherts of the various formations can be seen both at a macroscopic and petrographic level.
+Given the formation settings, petrographically, all the cherts from the Algarve are fairly homogeneous - marine origin, in limestone or dolomitic limestone rocks, all formed during the Jurassic.
 The use of fossils for the identification of the cherts is also difficult since these are often not well preserved enough to allow the identification of species that may connect a group of cherts.
@@
 These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana, and Malhão - to be differentiated without the need for thin sections.
-This is especially important for archaeological collections, especially those which may be small, with small archaeological pieces, or for the study of older collections to which other means of analysis may not be available.
+An exception might be the distinction between the cherts from the West and Malhão - the reddish cherts from Malhão are visually indistinguishable from those from the West with a higher fossil content.
+The grey cherts of the Malhão Middle Jurassic formation do show a higher concentration of visible round fossils, however the distinction is only possible seen under the stereomicroscope and on a fresh surface, which might hamper the classification of archaeological materials.
+These distinctions are especially important for archaeological collections, especially those which may be small, with small artifacts, or for the study of older collections to which other (destructive) means of analysis may not be applied.
 
-This data seems to confirm the potential of a macroscopic analysis to study the cherts of the Algarve.
+```{r macrocomparison, echo=FALSE, fig.cap="Comparison between the main chert types of the different outcrops in the Algarve, organized by geological age.
+The cherts from the Upper Jurassic are from the east section of the Algarve.", out.width='100%'}
+
+knitr::include_graphics("../figures/macro_comparison.png")
+
+```
+
+These data seem to confirm the potential of a macroscopic analysis to study the cherts of the Algarve.
 Albeit applying different methodologies, such as petrographic analyses, to these cherts is a way of completing the petrographic study of a collection, reliably applying mostly a macroscopic analysis to the assemblages coming from southern Portugal helps tackle issues such as the destructiveness, costliness, and time consumption of some methods.
-The research was able to complete the collection, providing samples of the available cherts which may be compared with archaeological samples.
-However, there are some caveats to this type of comparative studies.
-Landscapes have changed through time, especially with the influence of modern societies.
+Our study was able to provide a more detailed reference collection for chert outcrops in the Algarve, which will allow to test models about raw material procurement and use in a multilayered archaeological site like Vale Boi.
+There are, however, some noteworthy caveats in this type of study.
+Landscapes have changed through time, both naturally and with the influence of modern society.
 House constructions, agricultural fields, and roads, for example, have modified the landscape, possibly altering the availability and visibility of raw materials.
-The existence of chert at specific points on the map may not necessarily mean that they do not exist elsewhere.
-However, identifying characteristics that are common to most, if not all cherts of a region, may be a good indication that some cherts are not local.
-
+Other natural processes, such as the development of biomantle or soil cover may also hamper raw material visibility in the landscape.
+Similarly, environmental changes may have had an impact on raw material availability, through its impact on surface processes which expose, erode and transport the raw materials (Pereira and Benedetti, 2013).
+As such, it is important to keep in mind that current raw material sources, and specifically chert ones which may be sutbly visible in the landscape, may not correspond to the sources which were available in the past.
 Another caveat regarding chert sources, especially in a geographic area like the Algarve, is the possibility of some outcrops being submerged.
-Previous studies have identified the existence of Jurassic outcrops with chert nodules on the west coast, submerged by water.
-In times when the sea level was similar to the current one, these outcrops would not have been accessible, even during low tide, the only possibility for their procurement in nodules dragged to the shore.
-However, during periods where the sea level was lower due to water freezing in the polar caps (during the LGM, for example), large portions of the coast would have been accessible.
-This raises the question of whether the chert reference collection is representative of all the variability of chert in the region, including the submerged coastal areas.
-For studies focusing on periods like the Proto-Solutrean and Solutrean in the Algarve, for example, where nearly 20 km of the shore was exposed at the peak of the LGM, there may have been the exploitation of unknown sources or cherts with more variability than that known from the recovered samples.
-Despite the caveat, this raises the possibility to understand whether this new portion of landmass altered the raw material procurement patterns of these groups, or added new resources which had been previously unavailable.
-Studies that compare Gravettian and Magdalenian assemblages to Proto-Solutrean and Solutrean assemblages, within one single site, may give new insights to this question.
+Previous studies have identified the existence of Jurassic lithologies on the west coast, submerged by water (Terrinha et al., 2006).
+These were mainly surveys done by oil companies which were able to obtain the submerged stratigraphies on the southwestern coast of Portugal and that revealed Jurassic limestones and dolomites, although the presence of chert is not described and remains unknown.
+Whether these submerged lithologies are different from the currently emerged ones is also uncertain.
+These studies also revealed that Jurassic lithologies were covered by more recent geological layers, including Pliocene and Pleistocene layers, thus forming before and/or during Paleolithic occupations of the territory (Terrinha et al., 2006).
+In times when the sea level was similar to the current one, the submerged lithologies would not have been accessible, even during low tide.
+However, during periods when the sea level was lower due to water freezing in the polar caps, as during the LGM for example, large portions of the coast that had been submersed would have been accessible.
+During this period, in Portugal, the coast would probably be close to the continental platform, \~120 m below the sea level [@dias1997; @dias2000].
+Especifically in western Algarve, the coastline may have been displaced \~10-15 km offshore from the present coastline (Bicho and Haws, 2008).
+Although currently unknown, it may be possible that during periods like the LGM, limestone and dolomite Jurassic lithologies, with chert nodules, might have been exposed and available.
+When studying chronologies characterized by cold and harsh climatic conditions with impacts on sea level changes and mixed with coastal uplift events, it is relevant to keep in mind that the chert variability present in the current coastline may not necessarily reflect the variability in the past.
+Despite these caveats, our data raises the possibility to understand whether this new portion of landmass altered the raw material procurement patterns of these groups, or added new resources which had been previously unavailable.
+Studies that compare Gravettian and Magdalenian assemblages (with higher mean sea-levels, and possibly even similar to current coastlines) to Proto-Solutrean and Solutrean assemblages (with LGM low mean sea-levels), within one single site, may give new insights into this question.
 
@@
 
-The research and results in this paper continued and completed an important study of chert raw materials in southern Portugal.
-The results show the presence of four different chert formations, dispersed in the western and eastern Algarve.
-Within formations, there is a lot of variability in the nodules and the outcrops.
-There are however identifiable macroscopic and petrographic differences between formations.
-** Although the petrographic analysis is essential to identify the fossils present in the chert, a macroscopic approach seems to be pertinent for a quick and inexpensive analysis to distinguish between cherts of the different formations.
+In this study we identified several sources of chert nodules in southern Portugal and characterized the regional cherts, which were of critical importance for hunter-gatherer communities during the Late Pleistocene.
+For this we applied a multilayered raw material analysis approach, composed of macroscopic and petrographic analyses.
+The results show the presence of four different chert formations, dispersed in western and eastern Algarve.
+Within formations, there is variability in the nodules and the outcrops.
+There are however identifiable macroscopic and petrographic differences between formations which allow their distinction.
+Although the petrographic analysis is essential to identify the fossils present in the chert, a macroscopic approach seems to be pertinent for a quick and inexpensive analysis to distinguish between cherts of the different formations.
+The presence of chert sources in the Algarve region, with distinguisheable characteristics between formations, which may be analysed preliminarily through macroscopic approaches, shows the potential for chert raw material studies of archaeological sites in this key area.
+Further steps in our study will include the use of the data gathered in the present study and the completed LusoLit lithotheque to study chert use from multilayered sites with Upper Paleolithic chronologies such as Vale Boi, and participate in the discussion of human adaptations throughout the Late Pleistocene.
+Furthermore, future approaches include the use of geochemical methods to further characterize these cherts and the integration of the resulting data in the online database.
 

### 2022-12-21 analysis_backup/supplementary-materials/analysis_datasets.Rmd
- Commit after hard reset

--- 2022-10-06 analysis/supplementary-materials/petrography_dataset.Rmd
@@
 ---
-title: "Petrography dataset"
+title: "Analysis datasets"
+author: "Belmiro et. al"
 output: pdf_document
+classoption: landscape
 ---
@@
 
+knitr::opts_chunk$set(echo = FALSE, message = FALSE)
+
+```
+
+```{r callpackages}
+
+library(readr)
 library(knitr)
 library(kableExtra)
+
+```
+
+```{r calldb}
+
+macro_dataset <- read_csv("../data/supplementary_data/macroscopic_dataset.csv")
+
+knitr::kable(head(macro_dataset [1:3], 25), booktabs = TRUE, longtable = TRUE, caption = "Dataset for macroscopic description of the geological samples.") %>% 
+  column_spec(1, width = "4cm") %>%
+  column_spec(2, width = "3cm") %>%
+  column_spec(3, width = "12cm") %>% 
+  kable_styling(font_size = 9) %>% 
+  kable_styling(latex_options = c("repeat_header","hold_position"))
 
@@
   kable_styling(latex_options = "scale_down") %>% 
-  kableExtra::landscape() %>% 
   column_spec(2, width = "4 cm") %>% 

### 2022-12-22 analysis/paper/paper.Rmd
- Added PDF of SOM - Abstract and keywords

@@
 abstract: |
-  Text of abstract
+  Southwestern Iberia has played a key role in Middle to Upper Paleolithic transitions and Human occupations.
+It may have allowed the establishment of eco-cultural niches, through the continuous availability of resources throughout the Late Pleistocene, such as chert raw materials.
+Chert procurement studies in this region have received increasing attention in the past two decades, especially those related to multi-component sites such as Vale Boi (Algarve, Portugal).
+However, these studies have focused on the western portion of the region and used macroscopic descriptions.
+For further studies regarding Human mobility and raw material procurement and use at Vale Boi, a complete reference collection is necessary.
+This paper focuses on the location, diversity, and availability of chert raw materials in Southern Portugal (Algarve).
+Through macroscopic and petrographic approaches, we characterize the geological cherts to build a frame of reference for raw material studies in the region.
+Our results show that there are three main chert formations within Southern Portugal.
+Despite the variability within formations, the cherts show sufficient differences both at a macroscopic and petrographic level to allow source attribution.
+These results provide the starting ground for chert raw material studies in archaeological assemblages, in eco-cultural niches such as the archaeological site of Vale Boi.
+
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  Chert raw material; Petrography; Southern Portugal; Provenance studies
 ---

### 2022-12-22 analysis/paper/paper.Rmd
- Acknowledgements

@@
 
+We would like the thank the entities that funded the current research: Fundação para a Ciência e a Technologia (FCT) and ICArEHB (Interdisciplinary Center for Archaeology and the Evolution of Human Behavior).
+Furthermore, we want to thank Dr.
+Telmo Pereira and Dr.
+Carlos Ribeiro for allowing us to consult their previous research and for their support.
+Finally, we'd like to thank Jack Acres for the IT support essential for the creation of the online LusoLit database.
+
 <!-- The following line inserts a page break  -->

### 2022-12-22 README.Rmd
Some updates
- Read me file - Institution names

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia."
 ```
 
-# article1
+# Creating frame of reference for chert exploitation
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/jbelmiro/FramesReferenceChert/master?urlpath=rstudio)
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Belmiro, J., Cascalheira, J., Terradas, X., (-). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Belmiro, J., Cascalheira, J., Terradas, X., (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 

### 2022-12-22 analysis/paper/paper.Rmd
- Read me file - Institution names

@@
 institute:
-  - ICArEHB: Universidade do Algarve
-  - CSIC: Bleh
+  - ICArEHB: Interdisciplinary Center for Archaeology and the Evolution of Human Behavior, University of Algarve
+  - CSIC: Consejo Superior de Investigaciones Científicas
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2022-12-22 analysis/paper/paper.Rmd
JC reviews

--- 2022-12-22 analysis/paper/paper.Rmd
@@
     correspondence: true
+  - name: Xavier Terradas
+    email: 
+    institute: CSIC
+    correspondence: false
   - name: João Cascalheira
@@
     correspondence: false
-  - name: Xavier Terradas
-    email: 
-    institute: CSIC
-    correspondence: false
 institute:
-  - ICArEHB: Universidade do Algarve
-  - CSIC: Bleh
+  - ICArEHB: Interdisciplinary Center for Archaeology and the Evolution of Human Behavior, University of Algarve
+  - CSIC: Consejo Superior de Investigaciones Científicas
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 abstract: |
-  Southwestern Iberia has played a key role in Middle to Upper Paleolithic transitions and Human occupations.
-It may have allowed the establishment of eco-cultural niches, through the continuous availability of resources throughout the Late Pleistocene, such as chert raw materials.
-Chert procurement studies in this region have received increasing attention in the past two decades, especially those related to multi-component sites such as Vale Boi (Algarve, Portugal).
-However, these studies have focused on the western portion of the region and used macroscopic descriptions.
-For further studies regarding Human mobility and raw material procurement and use at Vale Boi, a complete reference collection is necessary.
-This paper focuses on the location, diversity, and availability of chert raw materials in Southern Portugal (Algarve).
-Through macroscopic and petrographic approaches, we characterize the geological cherts to build a frame of reference for raw material studies in the region.
-Our results show that there are three main chert formations within Southern Portugal.
-Despite the variability within formations, the cherts show sufficient differences both at a macroscopic and petrographic level to allow source attribution.
-These results provide the starting ground for chert raw material studies in archaeological assemblages, in eco-cultural niches such as the archaeological site of Vale Boi.
+  Southwestern Iberia has played a key role in characterizing Late Pleistocene human ecodynamics.
+Among other aspects of human behavior, chert procurement and use studies in this region have received increasing attention in the past two decades, especially focusing on the sites showing repeated human occupation, such as the case of Vale Boi (Southern Portugal).
+However, these studies have been very limited in their geographical scope, and mostly focused on brief macroscopic descriptions of the raw materials.
+To further our knowledge of the relationship between regional availability of raw materials and its impact on human adaptations and mobility, a more detailed approach to characterizing geological sources is needed.
+This paper characterizes chert raw materials location, diversity, and availability in a geologically well-defined region of southern Portugal - the Algarve.
+Through macroscopic and petrographic approaches, we provide a detailed characterization of geological chert sources to build a frame of reference for chert exploitation in the region.
+Our results show that there are three main chert formations in Algarve, and that despite the within-source variability, sufficient differences at macroscopic and petrographic levels are present to allow clear source attribution.
+These results provide a baseline for raw material studies in archaeological assemblages across southwestern Iberia, that will be essencial to further characterize the dynamics of human behavior in some of the most important eco-cultural niches.
 
@@
 The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c.
-37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022] .
+37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022].
 This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018].
@@
 Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources.
-This continuous use would then create long-term regional adaptative structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches.
+This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches.
 In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012].
@@
 One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi.
-This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of Iberia [@bicho2013; @cascalheira2013; @cascalheira2017].
+This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of southern Iberia [@bicho2013; @cascalheira2013; @cascalheira2017].
 Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017].
 Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials.
-The maintenancee of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016].
+The maintenance of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016].
 The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012].
-A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on macroscopic preliminary results or the differentiation between possibly local and non-local sorts of raw materials.
+A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on brief macroscopic results or the differentiation between possibly local and non-local sorts of raw materials.
 Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003].
@@
 Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999].
-In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like Vale Boi, it is necessary to know the landscape and the raw material sources available in the territory.
+In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like what is proposed for southwesternmost Iberia, it is necessary to know the landscape and the raw material sources available in the territory.
 To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage.
 This is true for most of the informal models which try to understand why raw materials change in the archaeological record.
-Furthermore, formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015].
+Formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015].
 In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022].
 In southern Portugal previous efforts have been made to create such a database.
-The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from Vale Boi based on macroscopic approaches.
+The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from some Late Pleistocene sites based on macroscopic approaches.
 The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region.
@@
 Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages.
-Furthermore, a large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists.
+A large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists.
 As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation.
@@
 This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts.
-The data presented here is integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.
+The data presented here is also fully integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.
 
@@
 Although the petrographic analysis is essential to identify the fossils present in the chert, a macroscopic approach seems to be pertinent for a quick and inexpensive analysis to distinguish between cherts of the different formations.
-The presence of chert sources in the Algarve region, with distinguisheable characteristics between formations, which may be analysed preliminarily through macroscopic approaches, shows the potential for chert raw material studies of archaeological sites in this key area.
+The presence of chert sources in the Algarve region, with distinguishable characteristics between formations, which may be analysed preliminarily through macroscopic approaches, shows the potential for chert raw material studies of archaeological sites in this key area.
 Further steps in our study will include the use of the data gathered in the present study and the completed LusoLit lithotheque to study chert use from multilayered sites with Upper Paleolithic chronologies such as Vale Boi, and participate in the discussion of human adaptations throughout the Late Pleistocene.
@@
 Carlos Ribeiro for allowing us to consult their previous research and for their support.
-Finally, we'd like to thank Jack Acres for the IT support essential for the creation of the online LusoLit database.
+We'd like to thank Jack Acres for the IT support essential for the creation of the online LusoLit database.
+João Cascalheira is funded by Portuguese Ministry for Science and Technology, contract reference DL57/2016/CP1361/CT0026.
+Current work on chert identification in Algarve is being supported by the European Research Council through the FINISTERRA project (101045506-FINISTERRA-ERC-2021-COG) attributed to João Cascalheira.
 

### 2022-12-22 analysis/paper/paper.Rmd
- Pre-knit corrections

@@
 Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999].
-In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like what is proposed for southwesternmost Iberia, it is necessary to know the landscape and the raw material sources available in the territory.
-To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage.
+In order to explore the aforementioned theories and models in an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like what is proposed for southwesternmost Iberia, it is necessary to know the landscape and the raw material sources available in the territory.
+To distinguish between local and non-local raw materials, first, it is necessary to understand the characteristics of the local sources and establish a comparative database that can be used for the analysis of an assemblage.
 This is true for most of the informal models which try to understand why raw materials change in the archaeological record.
@@
 In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022].
-In southern Portugal previous efforts have been made to create such a database.
+In southern Portugal, previous efforts have been made to create such a database.
 The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from some Late Pleistocene sites based on macroscopic approaches.
@@
 A large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists.
-As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation.
-Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity and availability of chert raw materials in the southernmost region of Portugal -- the Algarve.
-Our main goal is to establish a reference for chert raw materials in an understudied region in what regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere.
-This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts.
-The data presented here is also fully integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.
+As such, further analyses of the overall variability, location, and availability of chert nodules in the Algarve are necessary to start testing behavioral models of land-use and abiotic resource exploitation.
+Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity, and availability of chert raw materials in the southernmost region of Portugal -- the Algarve.
+Our main goal is to establish a reference for chert raw materials in an understudied region in regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere.
+This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterization of regional cherts.
+The data presented here is also fully integrated into an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.
 
@@
 The geological samples used in this study were obtained during fieldwork, between August 2021 and June 2022.
-The locations prospected were chosen after reviewing previously known research, which included preliminary raw materials studies in the region [@verissimo2004; @pereira2016], geological scientific papers and theses focusing on the Algarve basin and concerning chert outcrops [@rocha1976; @marques1983; @ramalho1985; @ribeiro2005], and geological maps, which signaled the presence of chert nodules within the outcrops[@rocha1979; @rocha1983; @rocha1989; @oliveira1984; @oliveira1992; @manuppela1987; @manuppella2007].
+The prospected locations were chosen after reviewing previously known research, which included preliminary raw materials studies in the region [@verissimo2004; @pereira2016], geological scientific papers and theses focusing on the Algarve basin and concerning chert outcrops [@rocha1976; @marques1983; @ramalho1985; @ribeiro2005], and geological maps, which signaled the presence of chert nodules within the outcrops[@rocha1979; @rocha1983; @rocha1989; @oliveira1984; @oliveira1992; @manuppela1987; @manuppella2007].
 Unpublished data and coordinates for unsurveyed locations with potential for chert-bearing outcrops gathered during the organization of the LusoLit lithotheque were also checked.
@@
 First, the samples were analyzed macroscopically following a pre-established dataset.
-The variables were defined based on specialized literature [@luedtke1992; @crandell2005; @bressy2002], and the dataset with variable description as well as the sample characterizations can be found in the SOM.
+The variables were defined based on specialized literature [@luedtke1992; @crandell2005; @bressy2002], and the dataset with the variable descriptions as well as the sample characterizations can be found in the SOM.
 A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with resource to a Nikon SMZ25 stereomicroscope, focusing primarily on inclusions and fossil content.
@@
 The complete R code used for all the analysis and visualizations contained in this paper is available at our online research compendium (DOI: 10.17605/OSF.IO/FP7TA).
-To produce those files, we followed the procedures described by Marwick et al.
-(2017) for the creation of research compendiums to enhance the reproducibility of research.
+To produce those files, we followed the procedures described by [@Marwick2017] for the creation of research compendiums to enhance the reproducibility of research.
 The files provided contain all the raw data used in our analysis as well as a custom R project [@wickham2015] holding the code to produce all tables and figures.
@@
 The banks seem to be mainly dolomite or dolomitic limestones.
-The process of dolomitization seems to have affected the chert nodules, as they often present different levels of silicification from the peripheral areas of the nodule to the interior, which also affects the size and feel of the grain .
+The process of dolomitization seems to have affected the chert nodules, as they often present different levels of silicification from the peripheral areas of the nodule to the interior, which also affects the size and feel of the grain.
 In this case, the peripheral areas of the chert nodules are more dolomitized, with visible grain and distinctively rough to the touch, while the interior areas are more silicified and conversely finer and smoother.
@@
 Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits.
-The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts.
+The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts (\@ref(fig:mjplate)).
 In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel.
@@
 All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all are poorly preserved and replaced by chalcedony or quartz).
-These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (Fig.
-\*\*).
+These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (\@ref(fig:mjplate)).
 
@@
 
-The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (Fig: ).
+The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (\@ref(fig:sourcesmap)).
 Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock.
@@
 The cherts show little macroscopic variability between nodule and outcrop.
-They are grey/brown (with visible yellow inclusions).
+They are grey/brown (with visible yellow inclusions) (\@ref(fig:ujplate)).
 Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas.
@@
 They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals.
-Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (Fig.
-\*\*).
+Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (\@ref(fig:ujplate)).
 
@@
 However, a small number of outcrops described in regional geological maps were not identified or were not accessible.
-This includes four outcrops: 1) a Lower Jurassic outcrop with chert micronodules which was identified through a geological profile (Oliveira, 1992); 2) a Middle Jurassic outcrop located at the easternmost section of the Algarve with partially dolomitized clasts and chert nodules (Oliveira, 1992); 3) several unprospected locations from previous geoarchaeological works, all related to the Peral Anticline and reaching from central to eastern Algarve; 4) a possible chert exploitation archaeological site located on top of a chert source (Monte do Cerro), in eastern Algarve which was not identified and is possibly inaccessible (Bicho et al., 2003).
+This includes four outcrops: 1) a Lower Jurassic outcrop with chert micronodules which was identified through a geological profile [@oliveira1992]; 2) a Middle Jurassic outcrop located at the easternmost section of the Algarve with partially dolomitized clasts and chert nodules [@oliveira1992]; 3) several unprospected locations from previous geoarchaeological works, all related to the Peral Anticline and reaching from central to eastern Algarve; 4) a possible chert exploitation archaeological site located on top of a chert source (Monte do Cerro), in eastern Algarve which was not identified and is possibly inaccessible [@bicho2003].
 All of these unidentifiable outcrops are located in the central or eastern portion of the Algarve.
@@
 The size, frequency, and preservation state of the fossils seem to be, then, one of the defining criteria for discerning cherts from different formations, and thus, different geographic areas.
-These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana, and Malhão - to be differentiated without the need for thin sections.
+These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana, and Malhão - to be differentiated without the need for thin sections (\@ref(fig:macrocomparison)).
 An exception might be the distinction between the cherts from the West and Malhão - the reddish cherts from Malhão are visually indistinguishable from those from the West with a higher fossil content.
@@
 Other natural processes, such as the development of biomantle or soil cover may also hamper raw material visibility in the landscape.
-Similarly, environmental changes may have had an impact on raw material availability, through its impact on surface processes which expose, erode and transport the raw materials (Pereira and Benedetti, 2013).
+Similarly, environmental changes may have had an impact on raw material availability, through its impact on surface processes which expose, erode and transport the raw materials [@pereira2013].
 As such, it is important to keep in mind that current raw material sources, and specifically chert ones which may be sutbly visible in the landscape, may not correspond to the sources which were available in the past.
 Another caveat regarding chert sources, especially in a geographic area like the Algarve, is the possibility of some outcrops being submerged.
-Previous studies have identified the existence of Jurassic lithologies on the west coast, submerged by water (Terrinha et al., 2006).
+Previous studies have identified the existence of Jurassic lithologies on the west coast, submerged by water [@terrinha2006].
 These were mainly surveys done by oil companies which were able to obtain the submerged stratigraphies on the southwestern coast of Portugal and that revealed Jurassic limestones and dolomites, although the presence of chert is not described and remains unknown.
 Whether these submerged lithologies are different from the currently emerged ones is also uncertain.
-These studies also revealed that Jurassic lithologies were covered by more recent geological layers, including Pliocene and Pleistocene layers, thus forming before and/or during Paleolithic occupations of the territory (Terrinha et al., 2006).
+These studies also revealed that Jurassic lithologies were covered by more recent geological layers, including Pliocene and Pleistocene layers, thus forming before and/or during Paleolithic occupations of the territory [@terrinha2006].
 In times when the sea level was similar to the current one, the submerged lithologies would not have been accessible, even during low tide.
@@
 During this period, in Portugal, the coast would probably be close to the continental platform, \~120 m below the sea level [@dias1997; @dias2000].
-Especifically in western Algarve, the coastline may have been displaced \~10-15 km offshore from the present coastline (Bicho and Haws, 2008).
+Especifically in western Algarve, the coastline may have been displaced \~10-15 km offshore from the present coastline [@bicho2008].
 Although currently unknown, it may be possible that during periods like the LGM, limestone and dolomite Jurassic lithologies, with chert nodules, might have been exposed and available.

### 2022-12-22 analysis/paper/paper.Rmd
- Final corrections before submission

@@
       fig_caption: yes
+      global_numbering: true
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
@@
 To the west and south, it is bordered by the Atlantic ocean.
-It extends for \~130 km E-W and \~50 km N-S and is characterized by a variety of geomorphic sub-regions and geological units, that make this region a complex territory (\@ref(fig:geomap)).
+It extends for \~130 km E-W and \~50 km N-S and is characterized by a variety of geomorphic sub-regions and geological units, that make this region a complex territory (Fig.
+\@ref(fig:geomap)).
 On the north sector of the Algarve, the Serra Algarvia is characterized by a mountainous range with a dense hydrographic network, which separates the Algarve from Alentejo.
@@
 When chert nodules within one single outcrop showed macroscopic differences (such as differences in the color, texture, translucency, or cortex), samples of each different nodule were collected, to cover all chert variability within the outcrop.
-This variability was also recorded using a database (to distinguish between homogeneous or heterogeneous chert nodules within the outcrop) and through photography (\@ref(fig:fieldphoto)).
+This variability was also recorded using a database (to distinguish between homogeneous or heterogeneous chert nodules within the outcrop) and through photography (Fig.
+\@ref(fig:fieldphoto)).
 All samples were registered with resource to a free android app (Archaeosurvey) which was designed for archaeological surveys, and records site location and characteristics [@cascalheira2017a], further adapted for raw material source surveys [@abrunhosa2017].
@@
 Thin sections were produced from geological samples of all formations, focusing on obtaining petrographic data that reflected the macroscopic variability.
-In total, 30 thin sections were produced (\@ref(tab:petrolist)), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve; 3) 1 thin-section of a geological sample recovered from previous works, which was not identified during our survey.
+In total, 30 thin sections were produced (Tab.
+\@ref(tab:petrolist)), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve; 3) 1 thin-section of a geological sample recovered from previous works, which was not identified during our survey.
 Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced.
@@
 Eighteen outcrops (primary and secondary) were revisited or newly identified in the Algarve region.
-Nine are located in the westernmost territory and nine to the east (\@ref(fig:sourcesmap)).
-From these, 57 samples were recovered and analyzed, of which 19 are isolated finds or in secondary settings (\@ref(tab:outcropdata)).
+Nine are located in the westernmost territory and nine to the east (Fig.
+\@ref(fig:sourcesmap)).
+From these, 57 samples were recovered and analyzed, of which 19 are isolated finds or in secondary settings (Tab.
+\@ref(tab:outcropdata)).
 
@@
 On the westernmost areas of the Algarve, there are mainly cherts from two different formations: Lower Jurassic and Upper Jurassic.
-The latter can be found in primary deposition in a single known outcrop - Praia da Mareta -- or nearby, in secondary deposition settings (\@ref(fig:sourcesmap)).
+The latter can be found in primary deposition in a single known outcrop - Praia da Mareta -- or nearby, in secondary deposition settings (Fig.
+\@ref(fig:sourcesmap)).
 Lower Jurassic outcrops are more common and, for that reason, have been more studied [@ribeiro2005].
-These outcrops are heterogeneous, showing different geological characteristics and chert colors (\@ref(fig:ljmacro)).
+These outcrops are heterogeneous, showing different geological characteristics and chert colors (Fig.
+\@ref(fig:ljmacro)).
 The Lower Jurassic cherts can be grouped into three main macroscopic types based on color (individual Munsell Color Chart codes can be found in the macroscopic description analysis table) and presence of fossil content: multicolored, yellow, red or purple type (MC); single grey/brown type (SGB); multicolored, yellow, red or purple with fossils type (MCF).
@@
 In more than 50% of the samples, no fossils can be identified, as all fossils, albeit common to very frequent, are poorly preserved, filled with chalcedony or quartz, and without any identifiable morphology.
-Whenever identifiable, fossils present in the sample are Echinoderms, Radiolarians, Sponge spicules, and bivalve shells (\@ref(fig:ljwest)).
+Whenever identifiable, fossils present in the sample are Echinoderms, Radiolarians, Sponge spicules, and bivalve shells (Fig.
+\@ref(fig:ljwest)).
 
@@
 The only identifiable difference is the presence of calcispheres.
-All samples from the Mareta outcrops and PtA seen under the petrographic microscope showed the presence of abundant calcispheres (\@ref(fig:ujwest)), which is not always apparent with the stereomicroscope.
-Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic (\@ref(fig:ujwest)), which were uncommon and scattered at the top of the cliffs by the beach.
+All samples from the Mareta outcrops and PtA seen under the petrographic microscope showed the presence of abundant calcispheres (Fig.
+\@ref(fig:ujwest)), which is not always apparent with the stereomicroscope.
+Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic (Fig.
+\@ref(fig:ujwest)), which were uncommon and scattered at the top of the cliffs by the beach.
 
@@
 Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits.
-The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts (\@ref(fig:mjplate)).
+The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts (Fig.
+\@ref(fig:mjplate)).
 In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel.
@@
 All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all are poorly preserved and replaced by chalcedony or quartz).
-These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (\@ref(fig:mjplate)).
+These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (Fig.
+\@ref(fig:mjplate)).
 
@@
 
-The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (\@ref(fig:sourcesmap)).
+The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (Fig.
+\@ref(fig:sourcesmap)).
 Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock.
@@
 The cherts show little macroscopic variability between nodule and outcrop.
-They are grey/brown (with visible yellow inclusions) (\@ref(fig:ujplate)).
+They are grey/brown (with visible yellow inclusions) (Fig.
+\@ref(fig:ujplate)).
 Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas.
@@
 They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals.
-Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (\@ref(fig:ujplate)).
+Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (Fig.
+\@ref(fig:ujplate)).
 
@@
 The size, frequency, and preservation state of the fossils seem to be, then, one of the defining criteria for discerning cherts from different formations, and thus, different geographic areas.
-These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana, and Malhão - to be differentiated without the need for thin sections (\@ref(fig:macrocomparison)).
+These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations - West, Jordana, and Malhão - to be differentiated without the need for thin sections (Fig.
+\@ref(fig:macrocomparison)).
 An exception might be the distinction between the cherts from the West and Malhão - the reddish cherts from Malhão are visually indistinguishable from those from the West with a higher fossil content.
